---
title: "Data Wrangling with R"
subtitle: "MSDA - Summer Bootcamp 2026"
author: "KT Wong"
email: "kwanto@hku.hk"
affiliation: "Faculty of Social Sciences, HKU"

date: "`r format(Sys.time(), '%B %d, %Y')`"

abstract: |
    The materials in this topic are drawn from @QSS, @Wickham2023, @Wickham2019 and @Wickham2016 as well as other sources, including Princeton Sociology Methods Camp 2023. The materials are for educational purposes only.

bibliography: reference_bootcamp.bib

format:
  revealjs:
    theme: clean_42_rb_revised5.scss
    self-contained-math: true
    embed-resources: true
    #show-notes: false
    multiplex: true
    slide-number: c/t
    preview-links: auto
    margin: 0.03
    height: 1080
    width: 1920
    max-scale: 2
    smaller: false
    scrollable: true
    incremental: true
    template-partials:
     - title-slide.html
    controls: true
    touch: false
    code-fold: true
    code-overflow: scroll
    code-line-numbers: true

execute:
    cache: true

jupyter: python3
engine: knitr

knitr:
  opts_chunk:
    echo: true
    eval: false
    message: false
    warning: false
    error: false
    results: 'hide'
    fig.show: 'hide'
    fig-cap-location: bottom
    fig-asp: 0.618

lightbox: auto
---


```{r setup}
#| echo: false
#| eval: false
#| include: false

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(tidyverse)
```

## Introduction

+ Data wrangling is the process of transforming and cleaning data to make it suitable for analysis
    +   it is an essential step in the data analysis process
    +   it involves a variety of tasks, including:
        +   loading data
        +   exploring data
        +   subsetting data
        +   manipulating data
        +   recoding variables
        +   reshaping data
        

+ we use the `tidyverse` package -  a collection of packages for data manipulation and visualization
    + the core of tidyverse package for data manipulation is the `dplyr` package
    + also use the `tidyr` package for reshaping data later
    
## Loading Data

+   The first step in any data analysis is to load the data into R
    +   the data can be in a variety of formats
    +   the most common formats are CSV, Excel, and SPSS
    +   the `readr` package is part of the tidyverse and is used to read data into R


```{css echo=FALSE}
code.sourceCode {
  font-size: 168%;
  /* or try font-size: xx-large; */
}
```

. . .

```{r read_data}

#install.packages("tidyverse")
library(tidyverse)


# check your working directory
getwd()

## read data file

addh <- read_csv(
  "https://raw.githubusercontent.com/kwan-MSDA/Bootcamp_2024/main/dataset/addhealthfake.csv"
)
```

## Add health dataset

+ Here are some details about the [dataset](https://addhealth.cpc.unc.edu/data/#public-use)
    + The National Longitudinal Study of Adolescent to Adult Health (Add Health)
    + a longitudinal study of a nationally representative sample of adolescents in **grades 7-12** in the United States during the 1994-95 school year (Wave I)
    + The Add Health cohort has been followed into young adulthood with four in-home interviews for Wave I-IV by 2008
        + the Wave V conducted during 2016-2018 included a mixed-mode survey
    + respondents’ social, economic, psychological, and physical well-being
        + along with contextual data on the family, neighborhood, community, school, friendships, peer groups, and romantic relationships
    + study developmental trajectories of health and risk behaviors throughout the life course
                               
+ The dataset used here is a subset of the Add Health dataset
    + 3000 observations and 11 variables
    
<!--

## Explore data

+ The dataset used here is a subset of the Add Health dataset
    + 3000 observations and 11 variables
        
+   After the dataset is loaded in R, it is important to explore the data to understand its structure and content

+   The following steps are recommended to explore the data:
    +   check the data types of key variables
    +   check the dimensions of the data
    +   look at a few rows and variables

. . .

```{r explore_data}

class(addh$age)

class(addh$gender)

class(addh$love)
```

## Explore data

+   For more information on the dataset (some re-cap), 
    + summary(): numeric summaries
    + str(): data types and sample data
    + colnames() or names(): names of columns/variables
    + dim(): dimensions
    + View(): view all data in RStudio viewer
    + head(): top 6 rows
    + tail(): bottom 6 rows



## Explore data

```{r explore_data1}

str(addh)
```
. . .

```{r explore_data2}

head(addh, n = 5)
```

. . .

```{r explore_data2a}

tail(addh, n = 5)
```

. . .

```{r}

colnames(addh)
```


::: {.notes}

+ (in dplyr) slice_sample(): randomly select n rows

::: 


## Explore data

+   To get information about one variable, use the following functions:
    + table(): get a table summarizing counts
    + unique(): get the unique responses for a variable
    + sort(): sort the numerically (or alphabetically)
    + hist(): produce a histogram

. . .

```{r explore_data3}

table(addh$gender)
```


. . .


```{r explore_data4}

table(addh$age)
```

. . .

```{r explore_data5}

sort(unique(addh$age))
```

## Subset data

+ let us recap by using the base R subsetting syntax with [row index, column index]

. . .

```{r subset_data1}
#| eval: false

# get first column, rows 1 through 3
addh[1:3, 1]
```

+ **Exercise**: How would you subset the observation in the third row and the fifth column?

+ we can use `-` to exclude rows or columns

. . .

```{r subset_data2}
#| eval: false

# get everything besides first row
addh[-1, ]
```

## Exercises

+ Suppose that you want to know a few things from the dataset:
    + What’s the median income of this sample? What’s the mean age?
    + What are the answer choices for debt?
    + On average, do the young adults surveyed think money, no cheating, or love is more important in a relationship?
    

+ hint: for simple calculations, you can use the base R functions mean(), median(), and table()

. . .

```{r exercises}
#| eval: false

median(addh$income)

mean(addh$age)

unique(addh$debt)
```
. . . 

```{r exercises1}
#| eval: false

# as a precursor to the next section, we can use dplyr to do the same thing

library(dplyr)

addh %>%
  summarise(
    mean_money = mean(money),
    mean_nocheating = mean(nocheating),
    mean_love = mean(love)
  )
```

-->

## Data Manipulation

+ The `dplyr` package is part of the tidyverse and is used for data manipulation
    
+ dplyr functions include:
    + `filter()`: subset rows
    + `select()`: subset columns
    + `mutate()`: create new variables
    + `summarise()`: summarize data
    + `arrange()`: sort data
    + `group_by()`: group data

+ very important function: pipe operator `%>%` from the **magrittr** package
    + allows you to chain functions together

+ basic structure of the dplyr functions
    + function(dataframe, operation 1 to perform, opertaion 2 to perform,...)

## Data Manipulation

[*`dplyr` - select*]{.big}

+ it can be used with [operations](https://dplyr.tidyverse.org/reference/select.html), like
    + `starts_with()`
    + `ends_with()`
    + `contains()`
    + `matches()`
    + etc...

. . .

```{r dplyr_select}
#| eval: false

pay_variables <- select(addh, contains("pay"))

head(pay_variables, 5)
```


## Data Manipulation

[*`dplyr` - filter*]{.big}

+ filter rows based on conditions

. . .

```{r dplyr_filter1}
#| eval: false

nodebt_income <- filter(addh, debt == "nodebt" & income >= 10000)

nrow(nodebt_income)
```
. . .

```{r dplyr_filter2}
#| eval: false

nomissing_income <- filter(addh, !is.na(income)) # only keep obs that are NOT missing

nrow(nomissing_income)

nomissing_income2 <- drop_na(addh, income) # alternate function from tidyr

nrow(nomissing_income2)
```

::: {.notes}

#nomissing_income <- drop_na(addh, income) # alternate function from tidyr

:::


## Data Manipulation

[*`dplyr` - arrange*]{.big}

+ sort data based on one or more columns

+ task: find the two observations who think money is extremely important for a relationship (10 on **money**) but who pay for the fewest percentage of dates (**paypercent**)

. . .

```{r dplyr_arrange}
#| eval: false

addh %>%
  filter(money == 10) %>%
  arrange(paypercent) %>%
  head(2)
```

. . .

```{r}
#| eval: false

addh %>%
  filter(money == 10) %>%
  arrange(paypercent, gender)
```

## Data Manipulation

[*`dplyr` - mutate*]{.big}

+ create new variables added to the dataset

+ task: add a variable with the average rating for nocheating, money, and love’s importance for a relationship (sum divided by 3) and another variable that logs that rating

. . .

```{r dplyr_mutate}
#| eval: false

addh <- mutate(
  addh,
  rateavg = (love + money + nocheating) / 3,
  rateavglog = log(rateavg)
)

head(addh, 5)
```


+ Caution
    + using the same column name or same object name, you overwrite the original object or column


## Data Manipulation

[*`dplyr` - group_by and summarise*]{.big}

+ group data by one or more variables and then summarize the data according to the groups

+ task: find the average "not cheating importance" for different gender

. . .

```{r dplyr_groupby}
#| echo: true
#| eval: false

addh %>%
  group_by(gender) %>%
  summarize(mean_nocheating = mean(nocheating))
```


+   **summarise** has a number of operations for creating summary statistics
    + mean(), median(), min(), max(), sd(), n(), n_distinct(), first, last, etc...

## Exercises

1. Exercise one
    + the number of females and males by debt status
    + the percentage in each (debt x gender) category as a fraction of all observations
    + the number of distinct ratings of love’s importance in each of these debt x gender categories

. . .

```{r exercises2}
#| eval: false

addh %>%
  group_by(gender, debt) %>%
  summarize(percentage = n() / nrow(addh), n_distinct_love = n_distinct(love))
```

## Exercises

2. Exercise two
  + Group the data by gender and debt status first
    + Find the average rating of love, no cheating, and money’s importance for a relationship in each group 
    + Arrange the groups by their rating of money’s importance to a relationship from the highest to rating to the lowest rating

. . .

```{r exercises3}
#| eval: false

addh %>%
  group_by(gender, debt) %>%
  summarize(
    mean_love = mean(love),
    mean_nocheating = mean(nocheating),
    mean_money = mean(money)
  ) %>%
  arrange(desc(mean_money))
```

## Recoding Variables

+   Recoding variables is a common task in data analysis for social science research
    +   convert a variable from one format to another
    +   create a new variable based on the values of an existing variable (or of multiple existing variables)

+   some typical recoding tasks
    +   convert a continuous variable to a categorical variable
    +   convert a categorical variable to a continuous variable
    +   create categorical variables based on conditions

::: notes

+   Our focus
    + data types
    + logical statements
    + case_when() function

:::

<!--

## Recoding Variables

[*`changing data types`*]{.big}

+   use the `mutate()` function to change the data type of a variable
    +   `as.character()`
    +   `as.numeric()`
    +   `as.factor()`
    +   `as.integer()`
    +   `as.logical()`

. . .

```{r recode_data1}

addh2 <- addh %>%
  mutate(
    age = as.character(age),
    debt = as.factor(debt)
  )

head(addh2, 3)
```


## Recoding Variables

[*`create a vector`*]{.big}

+ use c to string together the elements

. . .

```{r create_vector}
#| eval: false

agevec <- c(18, 21, 23, 25, 27, 30)

agevec
```

. . .

```{r create_vector1}
#| echo: true
#| eval: false

class(agevec)
```

. . .

```{r create_vector2}
#| echo: true
#| eval: false

gendervec <- c("male", "female", "other", "female", "female", "male")

gendervec
```

. . .

```{r create_vector3}
#| eval: false

class(gendervec)
```


## Recoding Variables

[*`create a vector`*]{.big}

+ Elements in a vector need to be of the same type, otherwise, type coercion happens

. . .

```{r create_vector4}
#| echo: true
#| eval: false

c(28, "28", TRUE)
```

. . .

```{r create_vector5}
#| echo: true
#| eval: false

c(28, "28", TRUE) %>% class()
```

. . .

```{r create_vector6}
#| echo: true
#| eval: false

c(1, 2, 3, TRUE, FALSE)

c(1, 2, 3, TRUE, FALSE) %>% class()
```

## Recoding Variables

[*`data types`*]{.big}

+ convert from one type to another using the following functions:
    + `as.numeric()`
    + `as.character()`
    + `as.factor()`

. . .

```{r create_vector7}
#| echo: true
#| eval: false

as.character(agevec)
```

. . .

```{r create_vector8}
#| eval: false

as.numeric(gendervec)
```

## Recoding Variables

[*`data types`*]{.big}

+ Vectors can have a factor type
    + looks like a character vector
    + but is actually a number under the hood (“labelled data”)

. . .

```{r create_vector9}
#| eval: false

genderfactorvec <- factor(gendervec, levels = c("male", "female", "other"))

genderfactorvec
```
. . .

```{r create_vector10}
#| eval: false

class(genderfactorvec)
```

. . .

```{r create_vector11}
#| eval: false

as.numeric(genderfactorvec)
```

-->

## Recoding Variables

[*`create a vector`*]{.big}

+   there are functions to help you create the vector more efficiently:
    1. `rep`: repeat the same thing multiple times
    2. `seq`: create a sequence of numbers
    3. `paste`: stick together character and numeric info
    4. `sample`: for vectors where we want to randomly sample from some larger pool

. . .

```{r create_vector12}
#| eval: false

rep(1, 5)
```

. . .

```{r create_vector12a}
#| eval: false

seq(from = 1997, to = 2025, by = 5)
```

. . .

```{r create_vector13}
#| eval: false

paste("age", seq(from = 22, to = 30, by = 1), sep = "_")
```



## Recoding Variables

[*`from vectors to dataframes`*]{.big}

+ One way to create a dataframe
    + use `bind_cols()` to attach same-length vectors together as columns in a tibble
    + Vectors can be different types

. . .

```{r create_dataframe}
#| eval: false

bind_cols(age = agevec, gender = gendervec)
```

## Exercises

. . . 

```{r exercises4}

addh2 <- addh %>%
  mutate(gender = factor(gender, levels = c("male", "female")))

str(addh2$gender)

vec1 <- as.character(addh2$gender)
vec2 <- as.numeric(addh2$gender)
```

+ convert the variable `gender` in addh to a factor variable

+ what happens if you try to convert the variable to character by using as.character after the factor conversion

+ what happens if you try to convert the variable to number by using as.numeric after the factor conversion


## Recoding Variables

[*`logical statements`*]{.big}

+ Recap: the main logical operators used in R are:
    + $==$ (equal to)
    + $!=$ (not equal to)
    + $!$ (not)
    + $<$ (less than)
    + $<=$ (less than or equal to)
    + $>$ (greater than)
    + $>=$ (greater than or equal to)
    + $\&$ (and)
    + $|$ (or)


## Recoding Variables
### logical statements

+ logical statements are used to filter data, create new variables, and recode variables
    + ifelse() function
    

```{r logical_statements}
#| class: fragment

addh2 <- addh %>%
  mutate(
    money_over_love = ifelse(money > love, 1, 0),
    .after = id
  )

head(addh2, 5)
```



```{r logical_statements1}
#| class: fragment

addh3 <- addh %>%
  mutate(
    money_or_love = ifelse(
      money == love,
      "same",
      ifelse(love > money, "love greater", "money greater")
    ),
    .after = id
  )

head(addh3, 5)
```


## Recoding Variables

[*`logical statements`*]{.big}

+ use `case_when()` if there are 3 or more conditions for creating a variable

+ its syntax is the following:
    + case_when(logical condition ~ value assigned, logical condition 2 ~ value assigned. . . .default = value if does not fit other logical conditions)


```{r logical_statements2}
#| class: fragment

addh3 <- addh %>%
  mutate(
    money_or_love = case_when(
      money == love ~ "same",
      love > money ~ "love greater",
      TRUE ~ "money greater"
    ),
    .after = id
  )

head(addh3, 5)
```

 
## Exercises

+ create a new variable called `money_or_love` in the `addh` dataset
    + the variable should have the following categories:
       + “extreme” if person either codes love or money as 9 or 10
       + “lovegreater” if love > money
       + “same” if love == money
       + “moneygreater” if money > love
       + NA if none of the above

## Recoding Variables
### create a binary vector

+ create a new variable called `high_income` in the `addh` dataset
    + the variable should be 1 if income is above the 75th percentile and 0 otherwise
    + use the `quantile()` function to find the 75th percentile of income

```{r create_vector14}
#| class: fragment

income_75 <- quantile(addh$income)[4]

addh2 <- addh %>%
  mutate(high_income = ifelse(income > income_75, 1, 0))
```

+ create a new variable called `income_level` in the `addh2` dataset
    + the variable should have the following categories:
       + “low” if income is below the 25th percentile
       + “high” if income is above the 75th percentile
       + “medium” if income is between the 25th and 75th percentiles
       
```{r create_vector15}
#| class: fragment

income_25 <- quantile(addh$income)[2]

addh2 <- addh2 %>%
  mutate(
    income_level = case_when(
      income <= income_25 ~ "low",
      income >= income_75 ~ "high",
      .default = "medium"
    ),
    .after = id
  )

head(addh2, 5)
```

## Looping

+   Loops are used to repeat a block of code multiple times
    +   for loop
    +   while loop
    +   repeat loop
    +   break and next statements
    
+   Loops are useful for:
    +   automating repetitive tasks
    +   iterating over elements in a list or vector
    +   creating new variables or dataframes
    +   running simulations

::: {.notes}

+   Loops can be slow in R
    +   use vectorized functions whenever possible
    +   use the apply family of functions
    +   use the purrr package

:::


## For loops

+   The for loop is the most common type of loop in R
    +   it repeats a block of code a specified number of times
    +   it can go through every element of a vector
        + syntax: `for (i in vector) {code to execute}`  
    +   it can iterate through a set number of elements in a vector
        + syntax: `for (i in 1:length(vector)) {code to execute}`
        
+  The for loop is also useful for:
    +   creating new variables
    +   running simulations


::: {.notes}

+ Steps to turn into a for loop:
    1. Initialize a vector to store results- this time it will store an entire vector of results rather than one result. Can either do:
        1.1 Initialize a vector of a certain length: vec <- vector(length = desired length)
        1.2 Initialize an empty vector: vec <- c()
    2. Use the for statement to tell the loop what to iterate through.
    3. Copy and paste the code from the single-observation case into the "meat" part of the for loop sandwich
    4. For step three, make sure to add indexing where appropriate

:::

## For loops --- sample means

+   Let us start with a simple example of a for loop
    +   we will use the `addh` dataset to calculate the sample means of the variable `money`
    +   we will take 1000 samples of size 800 with replacement and calculate the mean of each sample
    
```{r for_loop}
#| class: fragment

set.seed(123456)

sample_means <- numeric(length = 1000)

for (i in seq_along(sample_means)) {
  samp <- sample(addh3$money, size = 800, replace = TRUE)
  sample_means[i] <- mean(samp)
}

mean(sample_means)
```

+ plot the distribution of sample means

```{r for_loop1}
#| class: fragment

library(tidyverse)

ggplot(as.data.frame(sample_means), aes(sample_means)) +
  #geom_histogram(bins=30)+
  geom_density() +
  geom_vline(
    xintercept = mean(sample_means),
    color = "red",
    linetype = "dashed"
  ) +
  labs(
    title = "Distribution of Sample Means",
    x = "Sample Means",
    y = "Frequency"
  ) +
  theme_minimal()
```

## Write a function

+   Functions are blocks of code that can be reused
    +   they take input arguments
    +   they return output
    +   they can be used in loops, apply functions, and other functions
    +   they can be used to create new variables, summarize data, and run simulations
    
+ the basic structure of a function is:
    + `function_name <- function(input arguments) {code to execute}`
    + `return(output)`


+ we now touch upon some basics

::: {.notes}

1. What problem is the function trying to solve?
2. What are the function’s arguments in this case? (as a function’s inputs)
3. What is the function doing with those arguments?
4. What does the function return? What class is it? (check it using R)


:::


## functions - basics

+   Let us start with a simple function, z-score of a variable



```{r function1}
#| class: fragment

zscore <- function(x) {
  zscore <- (x - mean(x)) / sd(x)
  return(zscore)
}

z_income <- zscore(addh$income)

head(z_income, 10)
```


+ more complicated example

```{r function2}
#| class: fragment

sample_means <- function(data, n, reps) {
  sample_means <- numeric(length = reps)

  for (i in seq_along(sample_means)) {
    samp <- sample(data, size = n, replace = TRUE)
    sample_means[i] <- mean(samp)
  }

  return(sample_means)
}
```


```{r function3}
#| class: fragment

age_sample_means <- sample_means(addh$age, 500, 1000)

ggplot(as.data.frame(age_sample_means), aes(age_sample_means)) +
  geom_density() +
  geom_vline(
    xintercept = mean(age_sample_means),
    color = "red",
    linetype = "dashed"
  ) +
  labs(
    title = "Distribution of Sample Means",
    x = "Sample Means of Age",
    y = "Frequency"
  ) +
  theme_bw()
```

## Exercises

+ create a function that incorporate both sample means functions and the density plot as above

+ plot the distribution of sample means for the variable `love` in the `addh` dataset using the function


## Lists and map functions from `purrr` library

+   Lists are a way to store multiple objects in R
    +   can store vectors, dataframes, and other lists
    +   can store objects of different classes


```{r lists}
#| class: fragment

list1 <- list(1, c(8, 9, 10, 11, 12), data.frame(x = 1:10, y = 11:20))

list1
```


## Lists and map functions from `purrr` library

+ The syntax is `map(mylist, myfunction, functionoptions)` and can change depending on the type of output for your analysis

+ calculate the length of each element in the list above



```{r map_function}
#| class: fragment

library(purrr)

map(list1, length)
```


+ examine the class of each element in the list above

. . .

```{r map_function1}

map(list1, class)
```

+ do the summary of each element in the list above

. . .

```{r map_function2}

map(list1, summary)
```

## Tidy approach to data

+   The tidy approach (@Wickham2014) to data is a way to organize data in a consistent format
    + Each variable is a column; each column is a variable
    + Each observation is a row; each row is an observation
    + Each value is a cell; each cell is a single value

+   The tidy approach is useful for:
    +   data manipulation
    +   data visualization
    +   data analysis
    +   data sharing

+   The tidy approach is implemented in the `tidyr` package

::: notes

+   The tidy approach is a way to organize data in a consistent format
    +   it is based on the idea that data should be organized in a way that makes it easy to manipulate, visualize, and analyze
    +   it is a way to make data more accessible and easier to work with


+  Five common problems for messy data
    +   Column headers are values, not variable names
    +   Multiple variables are stored in one column
    +   Variables are stored in both rows and columns
    +   Multiple types of observational units are stored in the same table
    +   A single observational unit is stored in multiple tables
    
:::

## Reshape the data

### let us start from a "real" example

```{r reshape_data}
#| class: fragment

library(tidyr)

sleep_wide <- tibble(
  name = c("KT", "Wincy", "Kelly"),
  day1 = c(8, 7, 6),
  day2 = c(4, 6, 5),
  day3 = c(5, 4, 4),
  day4 = c(7, 7, 6)
)

sleep_wide
```

```{r reshape_data1}
#| class: fragment

sleep_long <- sleep_wide %>%
  pivot_longer(cols = day1:day4, names_to = "day", values_to = "sleep_hours")

sleep_long
```

```{r reshape_data1a}
#| class: fragment

library(stringr)

sleep_long %>%
  mutate(day = stringr::str_sub(day, -1, -1))
```


::: {.notes}

+ separate(col = office_housing, into = c("office", "housing"), sep = "_")

+ sleep_united <- sleep_pls_unite %>% unite(col = "office", building, floor, sep = "")

:::



## Reshape the data

### from long to wide

```{r reshape_data2}
#| class: fragment

sleep_wide2 <- sleep_long %>%
  pivot_wider(names_from = day, values_from = sleep_hours)

sleep_wide2
```

## Execises
+   The data is in a wide format, with each day as a separate column
    +   how to reshape the data to have each day as a separate row?
    +   how to reshape the data to have each activity (sleep, play) as a separate column?
    +   how to reshape the data to have each activity (sleep, play) as a separate row?

```{r reshape_data3}
#| class: fragment

sleep_wide2 <- tibble(
  name = c(rep("KT", 2), rep("Wincy", 2), rep("Kelly", 2)),
  activity = rep(c("sleep", "play"), 3),
  day1 = c(8, 2, 7, 2, 5, 3),
  day2 = c(6, 1, 1, 3, 6, 2),
  day3 = c(5, 1, 4, 1, 4, 3),
  day4 = c(7, 2, 7, 3, 6, 2)
)

sleep_wide2
```

. . . 

+ how to express the data in a tidy format? 
    + i.e. name, day, sleep, play as columns

. . .

```{r reshape_data4}
#| class: fragment

sleep_tidy <- sleep_wide2 %>%
  pivot_longer(cols = day1:day4, names_to = "day", values_to = "hours") %>%
  mutate(day = stringr::str_sub(day, -1, -1)) %>%
  pivot_wider(names_from = activity, values_from = hours)

sleep_tidy
```

## Export data - Recap

+   The last step in data analysis is to export the data
    +   save the data in a format that can be shared with others
    +   save the data in a format that can be read by other software
    +   save the data in a format that can be used in other software
    
+ Export command depends on the type of file you are trying to write to
    + write.csv for CSV
    + write.xslx for Excel 
    + write.dta for Stata file
    + etc
    
+ By default, the new file will be saved in current working directory
    + If you want to save it elsewhere, need to specify the path
    
. . .

```{r export_data}
write_csv(sleep_tidy, "sleep_tidy.csv")

library(haven)
write_dta(sleep_tidy, "c:/Users/KT/Downloads/sleep_tidy.dta")
```

## Basic merge

+ The typical merge in R is `left_join`
    + keep all rows from “left” table even if observation doesn’t have matching row in “right” table
    + this will drop observations from the joining data when they are not matched to the "left" table

. . .

```{r merge_data}

sleep_tidy <- tibble(
  name = c("KT", "Wincy", "Kelly", "May", "Mary"),
  sleep = c(8, 7, 6, 5, 5),
  play = c(2, 2, 3, 3, 2)
)

sleep_tidy2 <- tibble(
  name = c("KT", "Wincy", "Kelly", "Peter", "Susan"),
  study = c(3, 4, 5, 2, 3),
  work = c(8, 10, 9, 6, 5)
)

sleep_tidy3 <- left_join(sleep_tidy, sleep_tidy2, by = "name")

sleep_tidy3
```

+ put different dataset on the left side of the join

. . .

```{r merge_data1}
#| echo: true

sleep_tidy4 <- left_join(sleep_tidy2, sleep_tidy, by = "name")

sleep_tidy4
```


## Merge data

+ **inner join** Only keep rows of the first `data.frame` that have corresponding records in the second `data.frame`
    + this will drop observations from both dataframes when they are not matched to each other

. . . 

```{r merge_data2}

sleep_tidy5 <- inner_join(
  sleep_tidy,
  sleep_tidy2,
  by = "name",
  suffix = c("_sleep", "_work")
)

sleep_tidy5
```

. . .

+ **full join** Keep all rows from both dataframes, filling in missing values with NAs


. . .

```{r merge_data3}
#| output-location: fragment

sleep_tidy6 <- full_join(sleep_tidy, sleep_tidy2, by = "name")

sleep_tidy6
```

## References {visibility="uncounted"}

::: {#refs}
:::


