---
title: "Probability and Distribution"
subtitle: "MSDA - Summer Bootcamp 2026"
author: "KT Wong"
email: "kwanto@hku.hk"
affiliation: "Faculty of Social Sciences, HKU"
date: "`r format(Sys.time(), '%B %d, %Y')`"
abstract: |
    The materials in this topic are drawn from @QSS, @Wickham2023, @Wickham2019 and @Wickham2016 as well as other sources, including Princeton Sociology Methods Camp 2023. The materials are for educational purposes only.

bibliography: reference_bootcamp.bib
format:
  revealjs:
    theme: clean_42_rb_revised5.scss
    self-contained-math: true
    embed-resources: true
    #show-notes: false
    multiplex: true
    slide-number: c/t
    preview-links: auto
    margin: 0.03
    height: 1080
    width: 1920
    max-scale: 2
    smaller: false
    scrollable: true
    incremental: true
    template-partials:
     - title-slide.html
    controls: true
    touch: false
    code-fold: true
    code-overflow: scroll
    code-line-numbers: true
    mathjax: true

execute:
    cache: true

jupyter: python3
engine: knitr

knitr:
  opts_chunk:
    echo: true
    eval: true
    message: false
    warning: false
    error: false
    results: 'markup'
    fig.show: 'hold'
    fig-cap-location: bottom
    fig-asp: 0.618

lightbox: auto
---


## Motivation

+ in many cases we must figure out ways to distinguish systematic patterns from noise

+ Probability is a set of mathematical tools that measure and model randomness in the world

+ in social sciences, we use probability to model the randomly determined nature of various realworld events as well as human behavior and beliefs
  + our task is to identify systematic patterns from apparently noisy data

+ we use probability to quantify the degree of uncertainty about empirical findings

::: notes

+ Quantifying the degree of statistical uncertainty about empirical findings

+ Noise, also known as random error, is the irrelevant variation that occurs in every realworld data set

+ Randomness does not necessarily imply complete unpredictability

:::

## Two schools of thought in probability

+ **Frequentist**
  + probability represents `the limit of the relative frequency`
    + the ratio between the number of times the event occurs and the number of trials, in repeated trials under the same conditions
  + **objective** interpretation of probability
  + assumes that all outcomes are equally likely

+ **Bayesian**
  + probability is a measure of one's **subjective belief** about the likelihood of event occurrence
  + named after a eighteenth century English mathematician and minister, Thomas Bayes

+ Probability was established as a mathematical theory by Russian mathematician **Andrey Kolmogorov** in the early twentieth century
    + Kolmogorov's axioms of probability provide a mathematical foundation for both frequentist and Bayesian interpretations

::: notes

+ under the Bayesian framework, probability simply becomes a tool to describe one’s belief system rather than real-world phenomena

+ **Frequentist**: probability is the long-run relative frequency of an event occurring in repeated trials
+ **Bayesian**: probability is a measure of one’s subjective belief about the likelihood of event occurrence
+ Kolmogorov's axioms of probability provide a mathematical foundation for both frequentist and Bayesian interpretations
+ Kolmogorov's axioms are a set of three rules that define the properties of probability measures
+ Kolmogorov's work established probability theory as a rigorous mathematical discipline, providing a framework for both frequentist and Bayesian approaches

:::
  
## Deinfition of Probability

+ The definition of probability involves three concepts
  1. **experiment**: an action or a set of actions that produce stochastic events of interest
  2. **sample space**: a set of all possible outcomes of the experiment, typically denoted by $\Omega$
  3. **event**: a subset of the sample space

+ An event represents any subset of sample space
  + it may include multiple outcomes
  + the entire sample space that contains all outcomes is also an event
  + an event occurs if the set that defines the event includes an actual outcome of the experiment
  
+ the probability of event $A$ occurring, denoted by $P(A)$, can be computed as 

. . .

$$
P(A) = \frac{\text{number of elements in } A}{\text{number of elements in } \Omega}
$$

::: notes

+ an experiment of tossing a fair coin three times

+ When each outcome is equally likely, in order to compute the probability of event A 
  + we need to count the number of elements in the event A, as well as the total number of elements in the sample space 
  
:::

## Axioms of Probability

+ The probability axioms are given by the three rules below
  1. The probability of any event A is nonnegative: $P(A) ≥ 0$
  2. The probability that one of the outcomes in the sample space occurs is `1`: $P(\Omega)= 1$
  3. If events *A* and *B* are `mutually exclusive`, then $P(A \textrm{ or } B) = P(A) + P(B)$

+ For any given events *A* and *B*, the addition rule is given by $P(A \textrm{ or } B) = P(A) + P(B) − P(A \textrm{ and } B)$

+ the complement of a set -  all elements in the sample space except those in the set
  + the complement of event *A* is denoted by $A^c$, and its probability is given by $P(A^c) = 1 - P(A)$
  + e.g the complement of the event ${A \textrm{ or } B}$ is ${A^c \textrm{ and } B^c}$  

+ For any given events **A** and **B**, the **law of total probability** is given by

. . .

$$
P(A) = P(A \textrm{ and } B) + P(A \textrm{ and } B^c) 
$$  
  
::: notes

+ The first two axioms together imply that probability ranges from 0 to 1

+ “Mutually exclusive” in the last axiom means that two events A and B do not share an outcome
  + Using the Venn diagram, mutually exclusive events imply two disjoint sets, which do not share any element

+ decompose the event ${A \textrm{ or } B$ into three mutually exclusive events
  + ${A \textrm{ and } B^c}$, ${A^c \textrm{ and } B}$, and ${A \textrm{ and } B}$
  
:::

## Permutations

+ **Permutation** is an arrangement of objects in a specific order

+ e.g. the number of ways to arrange three objects **A**, **B**, and **C** is $6$
  + $ABC, ACB, BAC, BCA, CAB, CBA$

+ the number of permutations of **n** objects is given by **n!** (**n** factorial)

+ the factorial of a non-negative integer $n$ is the product of all positive integers less than or equal to $n$
  + $n! = n \times (n-1) \times (n-2) \times ... \times 3 \times 2 \times 1$
  + e.g. $3! = 3 \times 2 \times 1 = 6$
  
+ the factorial of **0** is **defined** as **1**, i.e. $0! = 1$

+ The number of permutations when arranging $k$ objects out of $n$ unique objects is given by $_nP_k=\frac{n!}{(n-k)!}$
  
+ e.g. the number of ways in which you can arrange **4** cards out of **13** unique cards 
  + $_{13}P_4 = \frac{13!}{(13-4)!} = \frac{13!}{9!} = 13 \times 12 \times 11 \times 10 = 17160$


::: notes

+ Permutations refers to the number of ways in which objects can be arranged
  + the order of arrangement matters
  + e.g. the number of ways to arrange three objects A, B, and C is 6
    + ABC, ACB, BAC, BCA, CAB, CBA

:::

## Permutations - The birthday problem

+ the number of people one needs in order for the probability that at least two people have the same birthday to exceed $0.5$, assuming that each birthday is equally likely

+ let $k$ be the number of people

. . .

$$
\begin{align*}
P(\textrm{at least two people have the same birthday}) & \\
= 1 − P(\textrm{nobody has the same birthday})
\end{align*}
$$

. . .

$$
\begin{align*}
\implies P(\textrm{nobody has the same birthday}) & \\
= \frac{_{365}P_k}{365^k} = \frac{365!}{(365-k)! \cdot 365^k}
\end{align*}
$$

## Permutations - The birthday problem

$$
\begin{align*}
\implies \log P(\textrm{nobody has the same birthday}) & \\
= \log 365! - k \log 365 - \log(365-k)!
\end{align*}
$$


+ use R to compute the probability

```{r}
#| echo: True
#| class: fragment
#| code-summary: "Compute the probability of the birthday problem"

# create a new function
birthday <- function(k) {
  logdenom <- k * log(365) + lfactorial(365 - k)
  lognumer <- lfactorial(365)
  pr <- 1 - exp(lognumer - logdenom)
  pr
}
```

```{r}
#| echo: True
#| output-location: column-fragment
#| class: fragment
#| code-summary: "Plot the probability of the birthday problem"
#| fig-width: 6
#| fig-height: 4
#| out-height: 4in

library(tidyverse)

# create a data frame with k values from 1 to 100

k_values <- tibble(k = 1:100) %>%
  mutate(probability = birthday(k))

# plot the probability

ggplot(k_values, aes(x = k, y = probability)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(
    title = "Probability of at least two people having the same birthday",
    x = "Number of people (k)",
    y = "Probability"
  ) +
  theme_minimal() +
  scale_y_continuous(
    str_c(
      "Probability that at least two",
      "people have the same birthday",
      sep = "\n"
    ),
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.1)
  )
```

::: notes

+ when the number of people equals $23$, the probability of at least two people having the same birthday exceeds $0.5$
  + when the number of people is more than $50$, this probability is close to $1$

+  Birthday problem: the answer is only 23 people, which is much lower than what most people think
  + To solve this problem using permutations, first notice the following relationship:
    + $P(\textrm{at least two people have the same birthday}= 1 − P(\textrm{nobody has the same birthday})$
  + Let $k$ be the number of people
    + the number of ways in which $k$ unique birthdays can be arranged out of 365 days
      + $_{365}P_k = \frac{365!}{(365-k)!}$
    + the total number of elements in the sample space is equal to the total number of ways to select $k$ possibly non-unique birthdays out of 365 days
  + $P(\textrm{nobody has the same birthday})$= $\frac{\textrm{# of ways in which k unique birthdays can be arranged}}{\textrm{# of ways in which k possibly nonunique birthdays can be arranged}}$
    + the number of ways in which $k$ unique birthdays can be arranged is given by $_{365}P_k = \frac{365!}{(365-k)!}$
    + the total number of ways in which $k$ possibly nonunique birthdays can be arranged is given by $365^k$
      + this is because each person can have any of the 365 days as their birthday, and there are $k$ people
    + thus, the probability that nobody has the same birthday is given by
      + $P(\textrm{nobody has the same birthday}) = \frac{_{365}P_k}{365^k} = \frac{365!}{(365-k)! \cdot 365^k}$

:::


## Sampling with and without replacement

+ The Monte Carlo simulation method
  + a general class of stochastic methods that can be used to approximately solve analytical problems by randomly generating quantities of interest
  
+ For the birthday problem, we sample $k$ possibly nonunique birthdays out of $365$ days and check whether or not the sampled $k$ birthdays are all different
  + this is called **sampling with replacement**
  + for each of $k$ draws, every one of $365$ days is equally likely to be sampled regardless of which dates were sampled before

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-fold: true
#| code-summary: "Monte Carlo simulation for the birthday problem"

set.seed(123) # for reproducibility
k <- 23

n_trials <- 10000 # number of simulation trials
event <- 0

for (i in 1:n_trials) {
  # sample k birthdays with replacement
  birthdays <- sample(1:365, k, replace = TRUE)

  # check if there are any duplicates
  if (length(unique(birthdays)) < k) {
    event <- event + 1
  }
}

# compute the probability
probability <- event / n_trials

cat(
  "Estimated probability of at least two people having the same birthday (k =",
  k,
  "):",
  probability,
  "\n"
)
```

::: notes

+ repeating this sampling procedure many times, we compute the fraction of simulation trials where at least two birthdays are the same, and this fraction serves as an estimate of the corresponding probability

+ This difference (from 0.507) is called the Monte Carlo error, which is an inevitable consequence of the simulation approach
  + The size of the Monte Carlo error depends on the nature of the problem
  + It is difficult to eliminate such an error but possible to reduce it
  + To obtain a more accurate estimate, we increase the number of simulations

:::

## Combinations

+ the number of ways to choose $k$ distinct elements out of $n$ elements without regard to their order

+ two different permutations **AB** and **BA** represent an identical combination

+ e.g. choose $2$ distinct elements out of $3$ elements **A**, **B**, and **C**
  + the number of permutations is $_3P_2 = 6$
  + the number of combinations is $3$

+ The number of combinations when choosing $k$ distinct elements from $n$ elements is denoted by ${n \choose k}$

. . .

$$
{n \choose k} = \frac{n!}{k!(n-k)!}
$$

+ e.g. Suppose that we are creating a committee of 5 out of 20 people (10 men and 10 women), assuming that each person is equally likely to be assigned to the committee
  + What is the probability that at least 2 women are on the committee? 

::: notes

+ P(at least 2 women are on the committee)= = 1 − P(no woman is on the committee) − P(exactly one woman is on the committee)

:::


## Conditonal Probability

+ Conditional Probability concerns how the probability of an event changes after we observe other events
  +  enables us to take into account observed evidence
  
+ The conditional probability of event **A** occurring given that event **B** occurred is defined as

. . .

$$
P(A|B)= \frac{P(A \textrm{ and B})}{P(B)}
$$

+ where $P(A \textrm{ and B})$ is the **joint probability** of events A and B occurring together, and $P(B)$ is the **marginal probability** of event B occurring

## Conditonal Probability

+ We can also express the joint probability of events **A** and **B** in terms of conditional probabilities

. . .

$$
P(A \textrm{ and } B) = P(A|B) \cdot P(B) = P(B|A) \cdot P(A)
$$

+ we can write the law of total probability in terms of conditional probabilities

. . .

$$
P(A) = P(A|B) \cdot P(B) + P(A|B^c) \cdot P(B^c)
$$

## Conditonal Probability - Example 1

+ consider two couples who are expecting twins
  + One couple had an ultrasound exam but the technician was only able to determine that one of the two children was a boy
  + The other couple did not find out the sex of their twins until the delivery, when they saw the first baby was a boy
  
+ Question: What is the probability that both babies are boys?


::: notes

+ the sample space = $\{GG, GB, BG, BB\}$
  + e.g. $GB$ means that the elder twin is a girl and the younger one is a boy

+ $P(BB| \textrm{ at least one is a boy }=\frac{P(BB)}{P(BB \text{ or } BG \textrm{ or } GB)}=\frac{1}{3}$

+ $P(BB | \textrm{ elder twin is a boy } = \frac{P(BB)}{P(BB \textrm{ or } BG)}=\frac{1}{2}$

:::

## Conditonal Probability - Example 2

+ a random sample of 10000 registered voters from Florida

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment

library(tidyverse)

# load the data
voters <- read_csv(
  "https://raw.githubusercontent.com/kwan-MSDA/soci7016_2025/refs/heads/main/data/FLVoters.csv"
)

dim(voters)

glimpse(voters)
```

+ county stands for county id of voter’s residence; VTD stands for voting district id of voter’s residence

+ remove those voters who contain a missing value

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment

voters <- voters %>%
  na.omit()

dim(voters)
```

::: notes

+ county - County id of voter’s residence
+ VTD - Voting district id of voter’s residence
+ age - Age
+ gender - Gender: m = male and f = female
+ race - Self-reported race

+ the dataset contains 9113 voters, with 5 variables
  + treat this sample of 9113 voters as a population of interest
  
:::


## Conditonal Probability - Example 2

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the marginal probability of race"

margin_race <-
  voters %>%
  count(race) %>%
  mutate(prop = n / sum(n))

margin_race
```

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the marginal probability of gender"

margin_gender <-
  voters %>%
  count(gender) %>%
  mutate(prop = n / sum(n))

margin_gender
```

:::

::: {.column width="50%"}
```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the conditional probability of race given female"

margin_race_f <- voters %>%
  filter(gender == "f") %>%
  count(race) %>%
  mutate(prop = n / sum(n))

cat(
  "Conditional probability of race given female voters:\n",
  paste(margin_race_f$race, ":", round(margin_race_f$prop, 4), collapse = "\n")
)
```

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the conditional probability of race given male"

margin_race_m <- voters %>%
  filter(gender == "m") %>%
  count(race) %>%
  mutate(prop = n / sum(n))

cat(
  "Conditional probability of race given male voters:\n",
  paste(margin_race_m$race, ":", round(margin_race_m$prop, 4), collapse = "\n")
)
```

:::

::::

::: notes

+ use the count() function to tally the number of observations for each value of the race variable
  + This is useful as it saves us from having to use the group() and ungroup() functions

:::


## Conditonal Probability - Example 2

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the joint probability of race and gender"

library(gt)
library(gtExtras)

joint_p <- voters %>%
  count(gender, race) %>%
  mutate(prop = n / sum(n))

joint_p %>%
  select(gender, race, prop) %>%
  arrange(gender, race) %>%
  gt() %>%
  tab_header(
    title = "Joint Probability of Race and Gender",
    subtitle = "Among Registered Florida Voters"
  )
```

:::

::: {.column width="50%"}
```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the probability of being an African American voter"

prob_black <- margin_race |>
  filter(race == "black") |>
  pull(prop)

cat(
  "Probability of being an African American voter:\n",
  round(prob_black, 4)
)
```


:::

::::

## The Marginal Probability of Race

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "the marginal probability of race"

joint_p_wider <- joint_p %>%
  select(-n) %>%
  pivot_wider(names_from = gender, values_from = prop) %>%
  mutate(total_prop = f + m)

joint_p_wider %>% 
  gt() %>%
  tab_header(
    title = "Marginal Probability of Race",
    subtitle = "Among Registered Florida Voters"
  ) %>% 
  gt_theme_538() %>%
  #gt_plt_bar(total_prop, palette = "blue") %>% 
  cols_label(
    race = "Race",
    f = "Female",
    m = "Male",
    total_prop = "Marginal Probability"
  ) %>% 
    text_transform(
      locations = cells_body(columns=race),
      fn = function(x) {
        case_when(
          x == "black" ~ "African American",
          x == "white" ~ "White",
          x == "hispanic" ~ "Hispanic",
          x == "asian" ~ "Asian",
          x == "other" ~ "Other",
          .default = x
        )
      }
    ) %>% 
      tab_style(
        style = cell_text(size = px(28)),
        locations = cells_body(columns = everything())
      ) %>% 
      tab_style(
        style = cell_text(size = px(28)),
        locations = cells_title(groups = "title")
      ) %>% 
      tab_style(
        style = cell_text(size = px(24)),
        locations = cells_title(groups = "subtitle")
      ) %>% 
      tab_style(
        style = cell_text(size = px(22)),
        locations = cells_column_labels(columns = everything())
      )

```

:::: {.columns}

::: {.column width="68%"}

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "the marginal probability of gender"

joint_p_wider <- joint_p %>%
  select(-n) %>%
  pivot_wider(names_from = race, values_from = prop) %>%
  mutate(total_prop = rowSums(across(where(is.numeric))))

joint_p_wider %>% 
  gt() %>%
  tab_header(
    title = "Marginal Probability of Gender",
    subtitle = "Among Registered Florida Voters"
  ) %>%
  gt_theme_538() %>%
  cols_label(
    gender = "Gender",
    black = "Black",
    white = "White",
    hispanic = "Hispanic",
    asian = "Asian",
    other = "Other",
    total_prop = "Marginal Probability"
  ) %>% 
    text_transform(
      locations = cells_body(columns=gender),
      fn = function(x) {
        case_when(
          x == "f" ~ "Female",
          x == "m" ~ "Male",
          .default = x
        )
      }
    ) %>% 
      tab_style(
        style = cell_text(size = px(22)),
        locations = cells_body(columns = everything())
      ) %>% 
      tab_style(
        style = cell_text(size = px(28)),
        locations = cells_title(groups = "title")
      ) %>% 
      tab_style(
        style = cell_text(size = px(24)),
        locations = cells_title(groups = "subtitle")
      ) %>% 
      tab_style(
        style = cell_text(size = px(22)),
        locations = cells_column_labels(columns = everything())
      )


```

:::

::: {.column width="32%"}

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "alternative method"

joint_p %>%
  group_by(gender) %>%
  summarise(prop = sum(prop)) %>% 
  gt() %>%
  tab_header(
    title = "Marginal Probability of Gender",
    subtitle = "Among Registered Florida Voters"
  ) %>%
  gt_theme_538() %>%
  cols_label(
    gender = "Gender",
    prop = "Marginal Probability"
  ) %>%
    text_transform(
      locations = cells_body(columns=gender),
      fn = function(x) {
        case_when(
          x == "f" ~ "Female",  
        x == "m" ~ "Male",
          .default = x
        )
      }
    ) %>% 
      tab_style(
        style = cell_text(size = px(28)),
        locations = cells_body(columns = everything())
      ) %>% 
      tab_style(
        style = cell_text(size = px(28)),
        locations = cells_title(groups = "title")
      ) %>% 
      tab_style(
        style = cell_text(size = px(24)),
        locations = cells_title(groups = "subtitle")
      ) %>% 
      tab_style(
        style = cell_text(size = px(22)),
        locations = cells_column_labels(columns = everything())
      )

```

:::

::::


::: notes

+ This joint probability gives $P$(black and female) = 0.07 and $P$(white and male) = 0.32
  + From this joint probability, we can compute the marginal and conditional probabilities

+ to obtain the marginal probability we apply the law of total probability 
  + $P(\textrm{black})=P(\textrm{black and female}) + P(\textrm{black and male})$
  
+ summing over gender columns for each racial row results in the marginal probability of race

+ To save us from having to type out the names of each race column in the mutate() function to create the total_prop variable
  + we use the rowSums() function with additional helper functions
  + The where() function allows us to select columns that match the specified criteria—here we select all the variables that are
numeric
  + The across() function applies a transformation over a series of columns
  + In short, the rowSums(across(where(is.numeric))) portion of the code tells R to sum the values of all the columns that are numeric
  
:::

## Conditonal Probability - Example 2

+ the conditional probability of being African American among female voters is calculated as

. . .

$$
P(\textrm{ black  } | \textrm{ female }) = \frac{P(\textrm{black and female})}{P(\textrm{ female })} = \frac{0.0744}{0.536}=0.139
$$

## Conditonal Probability - Example 2

+ a further example for conditional probability by creating $4$ age groups
  + **20 and below**, **21–40**, **41–60**, and **above 60**

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Create age groups"


voters <- voters %>%
  mutate(
    age_group = cut(
      age,
      breaks = c(0, 20, 40, 60, Inf),
      right = TRUE,
      labels = c("20 and below", "21-40", "41-60", "Above 60")
    )
  )

voters %>% 
  head() %>%
  gt() %>%
  tab_header(
    title = "Sample of Registered Florida Voters",
    subtitle = "With Age Groups"
  ) %>%
  gt_theme_538() %>%
  cols_label(
    county = "County",
    VTD = "Voting District",
    age = "Age",
    gender = "Gender",
    race = "Race",
    age_group = "Age Group"
  ) %>%
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_body(columns = everything())
    ) %>%
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_title(groups = "title")
    ) %>%
    tab_style(
      style = cell_text(size = px(24)),
      locations = cells_title(groups = "subtitle")
    ) %>%
    tab_style(
      style = cell_text(size = px(22)),
      locations = cells_column_labels(columns = everything())
    )
  
```

:::

::: {.column width="50%"}

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the marginal probability of age groups"


margin_age <- voters %>%
  count(age_group) %>%
  mutate(margin_age = n / sum(n)) %>%
  select(-n)

margin_age %>% 
  gt() %>%
  tab_header(
    title = "Marginal Probability of Age Groups",
    subtitle = "Among Registered Florida Voters"
  ) %>%
  gt_theme_538() %>%
  cols_label(
    age_group = "Age Group",
    margin_age = "Marginal Probability"
  ) %>%
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_body(columns = everything())
    ) %>%
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_title(groups = "title")
    ) %>%
    tab_style(
      style = cell_text(size = px(24)),
      locations = cells_title(groups = "subtitle")
    ) %>%
    tab_style(
      style = cell_text(size = px(22)),
      locations = cells_column_labels(columns = everything())
    )

```

:::

::::

## Conditonal Probability - Example 2

+ create a joint probability table across the three factors of race, gender, and age

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the joint probability of race, age group, and gender"


joint3 <- voters %>%
  count(race, age_group, gender) %>%
  mutate(prop = n / sum(n))

head(joint3) %>% 
  gt() %>%
  tab_header(
    title = "Joint Probability of Race, Age Group, and Gender",
    subtitle = "Among Registered Florida Voters"
  ) %>%
  gt_theme_538() %>%
  cols_label(
    race = "Race",
    age_group = "Age Group",
    gender = "Gender",
    prop = "Joint Probability"
  ) %>%
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_body(columns = everything())
    ) %>%
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_title(groups = "title")
    ) %>%
    tab_style(
      style = cell_text(size = px(24)),
      locations = cells_title(groups = "subtitle")
    ) %>%
    tab_style(
      style = cell_text(size = px(22)),
      locations = cells_column_labels(columns = everything())
    )
```

## Conditonal Probability - Example 2

+ the conditional probability of being African American and female given that a voter is above 60 

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "the conditional probability of race and gender given age group"

joint3 <- left_join(joint3, margin_age, by = "age_group") %>%
  mutate(cond_prob_age = prop / margin_age)

joint3 %>% 
  head(10) %>%
  gt() %>%
  tab_header(
    title = "Conditional Probability of Race and Gender Given Age Group",
    subtitle = "Among Registered Florida Voters"
  ) %>%
  gt_theme_538() %>%
  cols_label(
    race = "Race",
    age_group = "Age Group",
    gender = "Gender",
    cond_prob_age = "Conditional Probability"
  ) %>%
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_body(columns = everything())
    ) %>%
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_title(groups = "title")
    ) %>%
    tab_style(
      style = cell_text(size = px(24)),
      locations = cells_title(groups = "subtitle")
    ) %>%
    tab_style(
      style = cell_text(size = px(22)),
      locations = cells_column_labels(columns = everything())
    )

```

:::

:::: {.column width="50%"}

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment  
#| code-summary: "the conditional probability of being African American female given the voter is above 60"

filter(joint3, race == "black", gender == "f", age_group == "Above 60") %>%
  select(race, age_group, gender, cond_prob_age) %>% 
  gt() %>%
  tab_header(
    title = "Conditional Probability of Being African American and Female Given Age Group Above 60",
    subtitle = "Among Registered Florida Voters"
  ) %>%
  gt_theme_538() %>%
  cols_label(
    race = "Race",
    age_group = "Age Group",
    gender = "Gender",
    cond_prob_age = "Conditional Probability"                                   
  ) %>%
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_body(columns = everything())
    ) %>% 
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_title(groups = "title")
    ) %>%
    tab_style(
      style = cell_text(size = px(24)),
      locations = cells_title(groups = "subtitle")
    ) %>%
    tab_style(
      style = cell_text(size = px(22)),
      locations = cells_column_labels(columns = everything())
    ) 

```

:::

::::

## Conditonal Probability - Example 2

+ Q: the conditional probability of being African American given that a voter is female and above $60$
  + i.e. $P(\textrm{black| female and 60 years old })$
  + it can be computed by dividing the three-way joint probability $P(\textrm{ black and above 60 and female})$ by the two-way joint probability $P(\textrm{above 60 and female})$


```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the two-way joint probability of age group and gender"

joint2 <- voters %>%
  count(age_group, gender) %>%
  ungroup() %>%
  mutate(prob_age_gender = n / sum(n)) %>%
  select(-n)

```

+ merge with the 3-way joint probability and calculate conditional prob of race given age and gender


```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the conditional probability"
#| results: hide

joint3 <- left_join(joint3, joint2, by = c("age_group", "gender")) %>%
  mutate(con_prob_race = prop / prob_age_gender)

joint3 %>%
  gt() %>%
  tab_header(
    title = "Conditional Probability of Race Given Age Group and Gender",
    subtitle = "Among Registered Florida Voters"
  ) %>%
  gt_theme_538() %>%
  cols_label(
    race = "Race",
    age_group = "Age Group",
    gender = "Gender",
    con_prob_race = "Conditional Probability"
  ) %>%
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_body(columns = everything())
    ) %>%
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_title(groups = "title")
    ) %>%
    tab_style(
      style = cell_text(size = px(24)),
      locations = cells_title(groups = "subtitle")
    ) %>%
    tab_style(
      style = cell_text(size = px(22)),
      locations = cells_column_labels(columns = everything())
    )

```

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the conditional probability of being African American given female and above 60"

filter(joint3, gender == "f", age_group == "Above 60", race == "black") %>%
  select(con_prob_race) %>% 
  gt() %>%
  tab_header(
    title = "Conditional Probability of Being African American Given Female and Age Group Above 60",
    subtitle = "Among Registered Florida Voters"
  ) %>%
  gt_theme_538() %>%
  cols_label(
    con_prob_race = "Conditional Probability"
  ) %>%
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_body(columns = everything())
    ) %>% 
    tab_style(
      style = cell_text(size = px(28)),
      locations = cells_title(groups = "title")
    ) %>%
    tab_style(
      style = cell_text(size = px(24)),
      locations = cells_title(groups = "subtitle")
    ) %>%
    tab_style(
      style = cell_text(size = px(22)),
      locations = cells_column_labels(columns = everything())
    )

```       
       


::: notes

+ the cut() function is used to create a new variable age_group based on the age variable
  + breaks = c(0, 20, 40, 60, Inf) specifies the breakpoints for the age groups
  + right = TRUE indicates that the intervals are closed on the right side
  + labels = c("20 and below", "21-40", "41-60", "Above 60") specifies the labels for each age group

+ could do this using case_when

```{r}
#| echo: False
#| eval: False
#| include: False
#| output-location: fragment

voters <- voters %>%
  mutate(
    age_group = case_when(
      age <= 20 ~ "20 and below",
      age <= 40 ~ "21-40",
      age <= 60 ~ "41-60",
      TRUE ~ "Above 60"
    )
  )
```

:::


## Independence

+ the independence of two events implies that knowledge of one event does not give us any additional information about the occurrence of the other event

+ Two events **A** and **B** are independent if and only if

. . .

$$
P(A|B) = P(A) \textrm{ and } P(B|A) = P(B)
$$

+ Equivalently, two events **A** and **B** are independent if and only if

. . .

$$
P(A \textrm{ and } B) = P(A) \cdot P(B)
$$

## Independence

+ use our example to check whether the race and gender of registered Florida voters are independent
  + `crossing()` function: make every combination of the race and gender marginal probabilities
    + then merge that with the two-way joint probability table of race and gender, **joint_p**
    
:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Check independence of race and gender"

library(tidyr)

## rename needed variables
margin_race <- select(margin_race, race, prob_race = prop)

margin_gender <- select(margin_gender, gender, prob_gender = prop)

## combine the marginal probabilities and link to joint probabilities

race_gender_indep <- tidyr::crossing(margin_race, margin_gender) %>%
  left_join(
    select(joint_p, gender, race, prob_joint = prop),
    by = c("gender", "race")
  ) %>%
  ## calculate the independent prob comparison
  mutate(prob_indp = prob_race * prob_gender)

race_gender_indep %>% 
  gt() %>%
  tab_header(
    title = "Joint vs Independent Probabilities",
    subtitle = "Among Registered Florida Voters"
  ) %>%
  gt_theme_538() %>%
  cols_label(
    race = "Race",
    gender = "Gender",
    prob_joint = "Joint Probability",
    prob_indp = "Independent Probability"
  ) %>%
    tab_style(
      style = cell_text(size = px(18)),
      locations = cells_body(columns = everything())
    ) %>%
    tab_style(
      style = cell_text(size = px(22)),
      locations = cells_title(groups = "title")
    ) %>%
    tab_style(
      style = cell_text(size = px(20)),
      locations = cells_title(groups = "subtitle")
    ) %>%
    tab_style(
      style = cell_text(size = px(20)),
      locations = cells_column_labels(columns = everything())
    )


```

:::

::: {.column width="50%"}

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Plot joint vs independent probabilities for female"

ggplot(
  filter(race_gender_indep, gender == "f"),
  aes(x = prob_indp, y = prob_joint)
) +
  geom_point(shape = 1) +
  geom_abline(intercept = 0, slope = 1, color = "black", size = 0.5) +
  coord_fixed() +
  labs(
    title = "Joint vs Independent Probabilities",
    x = expression(P("race") * P("gender")),
    y = expression(P("race and gender"))
  )

```

:::

::::

## Independence

+ if we have three events **A**, **B**, and **C**, the joint independence among these events implies that the joint probability can be written as the product of marginal probabilities

. . .

$$
P(A \textrm{ and } B \textrm{ and } C) = P(A) \cdot P(B) \cdot P(C)
$$


+ The conditional independence of events **A** and **B** given event **C** implies that the joint probability of **A** and **B** given **C** is equal to the product of two conditional probabilities

. . .

$$
P(A \textrm{ and } B | C) = P(A | C) \cdot P(B | C)
$$

+ notice that joint independence implies pairwise independence


::: notes

+ This result can be obtained by applying the law of total probability:
$$
\begin{align*}
P(\textrm{A and B}) &= P(\textrm{A and B and C}) + P({\textrm{A and B and } C^c}) \\
&= P(A)P(B)P(C) + P(A)P(B)P(C^c) \\
&= P(A)P(B)\{P(C) + P(C^c)\} = P(A)P(B)
\end{align*}
$$

:::

## Independence

+ examine the joint independence among the sample of registered Florida voters

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Check joint independence of race, age group, and gender"

margin_age <- voters |>
  count(age_group) |>
  mutate(prob_age = n / sum(n)) |>
  select(age_group, prob_age)

## combine the marginal probabilities and link to joint probabilities
joint_indep <-
  tidyr::crossing(margin_age, margin_gender, margin_race) %>%
  mutate(indep_prob = prob_race * prob_age * prob_gender) %>%
  left_join(
    select(joint3, race, age_group, gender, prob_joint = prop),
    by = c("gender", "age_group", "race")
  )
```

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Plot joint vs independent probabilities for female above 60"

ggplot(
  filter(joint_indep, age_group == "Above 60", gender == "f"),
  aes(x = prob_joint, y = indep_prob)
) +
  geom_abline(intercept = 0, slope = 1, color = "black", size = 0.5) +
  geom_point(shape = 1) +
  coord_fixed() +
  labs(
    x = "P(race and above 60 and female)",
    y = "P(race) * P(above 60) * P(female)",
    title = "Joint independence"
  )
```


+  conditional prob of race and age given gender

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the conditional probability of race and age given gender"

cond_prob_gender <- left_join(
  select(joint3, race, age_group, gender, joint_prob = prop),
  margin_gender,
  by = c("gender")
) %>%
  mutate(cond_prob_gender = joint_prob / prob_gender)

```

+ conditional prob of race given gender

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the conditional probability of race given gender"

cond_prob_race_gender <-
  left_join(
    select(joint_p, race, gender, prob_race_gender = prop),
    margin_gender,
    by = "gender"
  ) %>%
  mutate(cond_prob_race_gender = prob_race_gender / prob_gender) %>%
  select(race, gender, cond_prob_race_gender)


```

## Independence

+ conditional prob of age given gender

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the conditional probability of age given gender"

cond_prob_age_gender <- left_join(
  select(joint2, age_group, gender, prob_age_gender),
  margin_gender,
  by = "gender"
) %>%
  mutate(cond_prob_age_gender = prob_age_gender / prob_gender) %>%
  select(age_group, gender, cond_prob_age_gender)
```

+ independent probability of race and age

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the independent probability of race and age given gender"

indep_cond_gender <- full_join(
  cond_prob_race_gender,
  cond_prob_age_gender,
  by = "gender"
) %>%
  mutate(indep_prob = cond_prob_race_gender * cond_prob_age_gender)

## merge the independent probability with the conditional probability

master <- left_join(
  cond_prob_gender,
  indep_cond_gender,
  by = c("gender", "age_group", "race")
)

```

+ plotting for women over 60

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Plot conditional vs independent probabilities for female above 60"

ggplot(
  filter(master, age_group == "Above 60", gender == "f"),
  aes(x = cond_prob_gender, y = indep_prob)
) +
  geom_abline(intercept = 0, slope = 1, color = "black", size = 0.5) +
  geom_point(shape = 1) +
  coord_fixed() +
  labs(
    x = "P(race and above 60 | female)",
    y = "P(race | female) * P(above 60 | female)",
    title = "Marginal independence"
  )
```


## Bayes' rule

+ Bayes’ rule shows how updating beliefs after observing some data

+ Bayes’ rule is a fundamental theorem of probability that relates the conditional and marginal probabilities of random events

+ Bayes’ rule is given by

. . .

$$
P(A|B) = \frac{P(B|A) \cdot P(A)}{P(B)}
$$

+ $P(A)$ is called the **prior probability** and reflects one’s initial belief about the likelihood of **event A** occurring
+ After observing the data, represented as **event B**, we update our belief and obtain $P(A | B)$, which is called the **posterior
probability**

## Bayes' rule

+ A well-known application of Bayes’ rule is the interpretation of medical diagnostic tests

+ Consider the following first-trimester screening test problem
  + A 35-year-old pregnant woman is told that 1 in 378 women of her age will have a baby with Down syndrome (DS)
  + A first-trimester ultrasound screening procedure indicates that she is in a high-risk category (HR)
  + Of 100 cases of DS, 86 mothers will receive a high-risk result (HR) and 14 cases of DS will be missed
  + Also, there is a 1 in 20 chance for a non-high-risk pregnancy to be diagnosed as high risk

+ Given the result of the screening procedure
  + what is the probability that her baby has DS?
  + What would the probability be if the result had been negative?


::: notes

+ P(DS | HR) = $\frac{P(HR | DS) P(DS)}{P(HR)}= \frac{P(HR | DS) P(DS)}{P(HR | DS)P(DS) + P(HR | \textrm{not DS})P(\textrm{not DS})}=\frac{0.86*\frac{1}{378}}{0.86*\frac{1}{378}+ 0.05*\frac{377}{378}}\approx 0.04$

+ P(DS | \textrm{not HR}) = $\frac{P(\textrm{not HR} | DS) P(DS)}{P(\textrm{not HR})}= \frac{P(\textrm{not HR} | DS) P(DS)}{P(\textrm{not HR} | DS)P(DS) + P(\textrm{not HR} | \textrm{not DS})P(\textrm{not DS})}=\frac{0.14*\frac{1}{378}}{0.14*\frac{1}{378}+ 0.95*\frac{377}{378}}\approx 0.0005$

:::

## Example --- Monty Hall Game

+ The Monty Hall problem is a famous probability puzzle based on a game show scenario
  + named after the original host of the American television game show **Let's Make a Deal**, Monty Hall
  
+ The game involves three doors, behind one of which is a car (the prize), and behind the other two are goats (not the prize)

+ The contestant chooses one door, then the host, who knows where the car is, opens one of the other two doors to reveal a goat
  + The contestant is then given the option to either stick with their original choice or switch to the other unopened door
  
. . .

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Simulate the Monty Hall game"

# create a function that simulates the game
sims <- 10000
doors <- c("goat", "goat", "car")
result.switch <- result.noswitch <- rep(NA, sims)

for (i in 1:sims) {
  ## randomly choose the initial door
  first <- sample(1:3, size = 1)
  result.noswitch[i] <- doors[first]
  remain <- doors[-first] # remaining two doors
  ## Monty chooses one door with a goat
  if (doors[first] == "car") {
    # two goats left
    monty <- sample(1:2, size = 1)
  } else {
    # one goat and one car left
    monty <- (1:2)[remain == "goat"]
  }
  result.switch[i] <- remain[-monty]
}

```

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the probability of winning the car when switching and not switching"

mean(result.noswitch == "car")

```

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the probability of winning the car when switching"

mean(result.switch == "car")

```

<!--

## Predicting race using surname and residence location

+ infer certain unknown attributes of individuals from their known characteristics
  + here, we consider the problem of predicting individual race using surname and residence location
  + Accurate prediction of individual race is useful when studying turnout rates among racial groups
 
+ The US Census Bureau releases a list of common surnames with their frequency

+ The data set includes a total of more than 150000 surnames that occurred at least 100 times
  + the census provides the relative frequencies of individual race within each surname using a six-category self-reported race measure
    + non-Hispanic White, non-Hispanic African American, non-Hispanic Asian and Pacific Islander, Hispanic origin, non-Hispanic American Indian and Alaskan Native, and non-Hispanic of two or more races
      + in the following analysis, we combine the last two categories into a single category of non-Hispanic others, so that we have five categories in total
      
::: notes

+ The self-reported information, which can be written as $P(race | surname)$, enables us to predict race given an individual’s surname

+ Note that $P(race)$, $P(race | surname)$, and $P(\textrm{race and surname})$ are examples of general ways to represent the marginal, conditional, and joint probabilities, respectively

:::

## Predicting race using surname and residence location

```{r}

library(tidyverse)

# load the data
cnames <- read_csv(
  "https://raw.githubusercontent.com/kwan-MSDA/soci7016_2025/refs/heads/main/data/cnames.csv"
)

glimpse(cnames)

cnames <- cnames %>%
  select(-1)

glimpse(cnames)
```

+ The total number of surnames contained in this data set is $151671$
  + For these names, the data set gives the probability of belonging to a particular racial group given a voter’s surname, i.e. $P(race | surname)$
  
## Predicting race using surname and residence location

+ adding a variable to the data that represents the most likely race for a given name
  + create a reshaped version of the data, then merge it back in with the original data
    + use the `pivot_longer()` function to reshape the data
    + within the `mutate()` function, we use the `str_replace` function to change the values of the `race_pred` variables

::: notes

+ In some Southern states, including Florida, voters are asked to self-report their race when registering

+ This makes the Florida data an ideal validation data set
  + If the accuracy of a prediction  method is empirically validated in Florida, we may use the method to predict individual race
in other states where such information is not available

:::

. . .

```{r}

most_likely_race <- cnames %>%
  select(-count) %>%
  pivot_longer(
    cols = -surname,
    names_to = "race_pred",
    values_to = "highest_pct"
  ) %>%
  mutate(race_pred = str_replace(race_pred, "pct", "")) %>%
  group_by(surname) %>%
  filter(row_number(desc(highest_pct)) == 1) %>%
  ungroup()

cnames <- full_join(cnames, most_likely_race, by = "surname")
```

+ use the `inner_join()` function to match names between the voter file and census name data
  + the `inner_join()` function retains only those rows where there is a match between the two data sets
  
. . .


```{r}

FLVotersJoin <- voters %>%
  inner_join(cnames, by = "surname")

dim(FLVotersJoin)

glimpse(FLVotersJoin)
```

## Predicting race using surname and residence location

+  there are some discrepancies between the values of race in the voters data and the values of the predicted race
  + The census data contain the categories "api" and "others", while the voter file data contain the categories "asian", "other", and
"native"

. . .

```{r}

unique(FLVotersJoin$race)

unique(FLVotersJoin$race_pred)
```
. . .

```{r}

FLVotersJoin <- FLVotersJoin %>%
  mutate(
    race = recode(race, "native" = "other"),
    race_pred = recode(race_pred, "api" = "asian", "others" = "other")
  )

unique(FLVotersJoin$race)

unique(FLVotersJoin$race_pred)
```

+ calculate the true positive rate for each racial group

. . .

```{r}

race_tp <- FLVotersJoin %>%
  group_by(race) %>%
  summarize(tp = mean(race == race_pred)) %>%
  arrange(desc(tp))

race_tp
```

::: notes

+ classification of Asian and African American voters is much worse
  + In particular, only 16% of African American voters are correctly classified as African American

+ The high true positive rate for Whites may simply arise from the fact that they far outnumber voters from other racial categories

:::

+ compute the false discovery rate for each racial group

. . .

```{r}

race_fp <- FLVotersJoin %>%
  group_by(race_pred) %>%
  summarize(fp = mean(race != race_pred)) %>%
  arrange(desc(fp))

race_fp
```


## Predicting race using surname and residence location

+ improve the above prediction by taking into account where voters live
  + there exists residential segregation based on race in the US
  + In the United States, voter files contain voters’ addresses
  
+ The FLCensus.csv data set contains the names of Florida counties and their corresponding county codes
  + The county codes are used to identify the counties in the voter file
  + The county names are used to identify the counties in the census data set
 
+ How does knowledge of residence location improve the prediction of individual race?
  + $P(\textrm{race | surname and residence})$

## Predicting race using surname and residence location

+ By applying Bayes' rule

. . .

$$
P(\textrm{race}| \textrm{surname and residence})=\frac{P(\textrm{surname| race and residence})P(\textrm{race|residence})}{P(\textrm{surname|residence})}
$$

+ $P(r\textrm{race | residence})$ is available from the Florida census data
  + however, the other two conditional probabilities, $P(\textrm{surname | race and residence})$ and $P(\textrm{surname |residence})$, are not directly available either in the census name data set or the Florida census data set

+ make an additional assumption that a voter’s surname and residence location are independent of each other given race

. . .

$$
P(\textrm{surname}| \textrm{race and residence})=\frac{P(\textrm{surname and residence|race})}{P(\textrm{residence|race})}=\frac{P(\textrm{surname})}{P(\textrm{race})}
$$

::: notes

+ This conditional independence assumption implies that once we know a voter’s race, their residence location does not give us any additional information about their surname


+ county -  County census id of voting district
+ VTD - Voting district census id (only unique within county)
+ total.pop - Total population of voting district
+ white - Proportion of non-Hispanic Whites in voting district
+ black - Proportion of non-Hispanic African Americans in voting district
+ api - Proportion of non-Hispanic Asian and Pacific Islanders in voting district
+ hispanic - Proportion of voters of Hispanic origin in voting district
+ others - Proportion of the other racial groups in voting district

+ Bayes' rule conditional on both B and C events
  + $P(A|B,C)=\frac{P(B|A \textrm{ and } C)P(A|C)}{P(B|C)}$

:::

## Predicting race using surname and residence location

$$
P(\textrm{surname|residence})=\sum_{\textrm{race}} P(\textrm{surname|race})P(\textrm{race|residence})
$$

+ $P(\textrm{surname|race})$ is not directly available from the dataset, but can be obtained by using Bayes' rule

. . .

$$
P(\textrm{surname|race})=\frac{P(\textrm{race|surname})P(\textrm{surname})}{P(\textrm{race})}
$$

+ $P(\textrm{race})$ can be computed from the Florida census data by using the law of total probability
  + $P(\textrm{race})=\sum\limits_{\textrm{residence}}P(\textrm{race|residence}) P(\textrm{residence})$

. . .

```{r}

library(tidyverse)

# load the data
FLCensus <- read_csv(
  "https://raw.githubusercontent.com/kwan-MSDA/soci7016_2025/refs/heads/main/data/FLCensusVTD.csv"
)

FLCensus <- FLCensus %>%
  rename("asian" = "api", "other" = "others")

race_prop <- FLCensus %>%
  select(total.pop, white, black, asian, hispanic, other) %>%
  pivot_longer(cols = -total.pop, names_to = "race", values_to = "pct") %>%
  group_by(race) %>%
  summarize(race_prop = weighted.mean(pct, w = total.pop)) %>%
  arrange(desc(race_prop))

race_prop
```

+ compute $P(\textrm{surname | race})$
  + use the `left_join()` function to add the race_prop data to the FLVotersJoin data
  + then record the total number of names in the cnames data for later use

. . .

```{r}

## merge the race probability with the existing data
FLVotersJoin <- left_join(FLVotersJoin, race_prop, by = "race")

## calculate prob of surname given race total number of individuals
total.count <- sum(cnames$count)

cnames_reshape <- cnames %>%
  select(-race_pred, -highest_pct) %>%
  pivot_longer(
    cols = starts_with("pct"),
    names_to = "race",
    values_to = "pct"
  ) %>%
  mutate(
    race = str_replace(race, "pct", ""),
    race = recode(race, "api" = "asian", "others" = "other"),
    race_surname = pct / 100
  ) %>%
  select(-pct) %>%
  left_join(race_prop, by = "race") %>%
  mutate(
    prob_surname_race = race_surname * (count / total.count) / race_prop
  ) %>%
  rename("posib_race" = race)

cnames_reshape
```


## Predicting race using surname and residence location

+ compute $P(\textrm{surname | residence})$

+ merge the Florida census data into the voter file data after reshaping to make sure it matches

+ merge in the reshaped cnames data

+ calculate the necessary quantity per race, then separately sum those quantities and merge the result back in

. . .

```{r}

## reshape the FL census data so we have race as a variable, then pct of pop
merge_temp <- FLCensus %>%
  pivot_longer(
    cols = c(white, black, hispanic, asian, other),
    names_to = "pop_race",
    values_to = "pop_pct"
  ) %>%
  inner_join(voters, by = c("county", "VTD")) %>%
  inner_join(cnames_reshape, by = c("surname", "pop_race" = "posib_race")) %>%
  mutate(race_residence = prob_surname_race * pop_pct)

## then calculate the summation
name_residence <- merge_temp %>%
  group_by(surname, county, VTD) %>%
  summarize(name_residence = sum(race_residence))
```

+ now we have all quantities of interest, can merge all together and calculate the predicted race

. . .

```{r}

FLVoters_full <- merge_temp %>%
  inner_join(name_residence, by = c("surname", "county", "VTD")) %>%
  mutate(pred_race = prob_surname_race * pop_pct / name_residence) %>%
  select(surname, pop_race, race, pred_race, county, VTD)
```

::: notes

+ What is crucial here is making sure we are merging with the correct fields specified to match between the data.

:::

+  see whether the highest predicted race estimate matches the actual race given by the respondent

. . .

```{r}

## filter to save the highest predicted race and see if it matches actual race

FL_updated <- FLVoters_full %>%
  ungroup() %>%
  group_by(surname, county, VTD) %>%
  ## select row per name with the largest percentage
  ## by keeping the first instance after arranging by percentage
  filter(row_number(desc(pred_race)) == 1) %>%
  ## ungroup to avoid errors later
  ungroup()

## calculate the new true positive rate
race_tp_new <- FL_updated %>%
  group_by(race) %>%
  summarize(tp = mean(race == pop_race)) %>%
  arrange(desc(tp))

race_tp_new
```

+  proportion of African Americans among those with surname "White"

. . .

```{r}

filter(cnames, surname == "WHITE") %>%
  select(pctblack) %>%
  pull()
```


+ predicted probability of being African American given residence location

. . .

```{r}

filter(FLVoters_full, surname == "WHITE", pop_race == "black") %>%
  select(pred_race) %>%
  summary()
```

+ compute the false positive rate for each race

. . .

```{r}

race_fp_new <- FL_updated %>%
  group_by(pop_race) %>%
  summarize(fp = mean(race != pop_race)) %>%
  arrange(desc(fp))

race_fp_new
```



::: notes

+ the true positive rate is much improved for Black American, jumping from 16% to 63%

+ Minor improvements are also made for Hispanic and Asian voters
  + suggesting that the location of a voter’s residence can be informative
  
+ For example, according to the census data, among people whose surname is “White,” 27% are African American

+ However, once we incorporate the location of their residence, the predicted probability of such individuals being African American ranges from 0% to 98%
  + This implies that we predict some voters to be highly likely African American and other voters to be highly likely non–African
American

+ the false positive rate for Whites is reduced, from 20% to 12%
  + This is in large part due to the fact that many of the African American voters who were incorrectly classified as Whites using surname alone are now predicted to be African American
  + We also see decreases in the false positive rate for African Americans, though it remains similar for the other races

:::

-->

## Random Variables and Probability Distributions

+ A random variable assigns a number to each event
  + two outcomes of a coin flip can be represented by a binary random variable where `1` indicates landing on heads and `0` denotes landing on tails
  
+ The values of random variables must represent mutually exclusive and exhaustive events

+ A random variable can be discrete or continuous
  + A discrete random variable can take on a finite or countably infinite number of values
    + e.g. the number of heads in 10 coin flips
  + A continuous random variable can take on any value within a given range
    + e.g. the height of a person
    
::: notes

+ consider a random variable that represents one’s racial group using five unique integers
  + black = 1, white = 2, hispanic = 3, asian = 4, and others = 5
  + someone who self-identifies as African American and White will be assigned the value of 5 instead of taking the values of 1 and 2 at the same time

+ Once we define a random variable, we can formalize a probability model using the distribution of the random variable

:::

## Bernoulli Distributions

+ A Bernoulli distribution is a discrete probability distribution for a random variable, $X$, that can take on only two possible outcomes
  + e.g. flipping a coin, where heads is represented by $1$ and tails by $0$
  + The Bernoulli distribution is parameterized by the probability of success, $p$, which is the probability of obtaining a $1$
  
+ The distribution of a discrete random variable can be characterized by the probability mass function (pmf)

+ The pmf $f(x)$ of a random variable $X$ is defined as the probability that the random variable takes a particular value, i.e. $f(x) = P(X = x)$

+ The pmf of a Bernoulli distribution is given by

. . .

$$
f(x) = \begin{cases}
p & \textrm{if } x = 1 \\
1-p & \textrm{if } x = 0 \\
0 & \textrm{otherwise }
\end{cases}
$$

## Bernoulli Distributions

+ the cumulative distribution function (CDF), $F(x)$ represents the cumulative probability that a random variable $X$ takes a value equal to or less than a specific value x, i.e., $F(x) = P(X ≤ x)$

+ the relationship between the PMF $f(x)$ and the CDF $F(x)$ for a discrete random variable can be written as

. . .

$$
F(x) = P(X ≤ x) =  \sum_{k\leq x} f(k)
$$

::: {.no-bullet}
where $k$ represents all values the random variable $X$ can take that are less than or equal to $x$
:::

+ The CDF of a Bernoulli distribution is given by

. . .

$$
F(x) = \begin{cases}
0 & \textrm{if } x < 0 \\
1-p & \textrm{if } 0 \leq x < 1 \\
1 & \textrm{if } x \geq 1
\end{cases}
$$

. . .


## Uniform Distribution

+ A uniform distribution is a continuous probability distribution where all outcomes are equally likely within a specified range, $[a,b]$
  + The uniform distribution is parameterized by two parameters, $a$ and $b$, which represent the lower and upper bounds of the distribution, respectively
  
+ use the (probability) density function (pdf) $f(x)$ to quantifies the likelihood that a continuous random variable $X$ will take a specific value $x$

+ The pdf of a uniform distribution is given by

. . .

$$
f(x) = \begin{cases}
\frac{1}{b-a} & \textrm{if } a \leq x \leq b \\
0 & \textrm{otherwise}
\end{cases}
$$

## Uniform Distribution

:::: {.columns}

::: {.column width="50%"}

the CDF of a uniform distribution is given by


$$
F(x) = \begin{cases}
0 & \textrm{if } x < a \\
\frac{x-a}{b-a} & \textrm{if } a \leq x < b \\
1 & \textrm{if } x \geq b
\end{cases}
$$

::: 

::: {.column width="50%"}

in general, the CDF of a continuous random variable $X$ can be expressed as

$$
F(x) = P(X ≤ x) = \int_{-\infty}^{x} f(t) dt
$$

:::

::::


```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the density of a uniform distribution"


dunif(0.5, min = 0, max = 1)

```

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the cumulative probability of a uniform distribution"

punif(1, min = -2, max = 2)

```


## Bernoulli and Uniform Distributions

+ We can use a uniform random variable to generate a Bernoulli random variable

+ in order to generate a Bernoulli random variable $Y$ with success probability $p$, we can first sample a uniform random variable $X$ and then set $Y = 1$ when $X$ is less than $p$ (similarly, set $Y = 0$ if $X ≥ p$) so that Y takes a value of 1 with probability $p$

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Generate a Bernoulli random variable from a uniform random variable"

set.seed(123)

sims <- 10000

p <- 0.5 # success probabilities

x <- runif(sims, min = 0, max = 1) # uniform [0,1]

head(x)
```

. . .

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Generate a Bernoulli random variable from a uniform random variable"

y <- as.integer(x <= p) # Bernoulli; turn TRUE/FALSE to 1/0

head(y)

```

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the mean of the Bernoulli random variable"

cat("Mean of Y:", mean(y))

```


## binomial distribution

+ a binomial random variable $X$ records the number of successes in a total of $n$ independent and identical trials with success probability $p$
  + e.g. the number of heads in $10$ coin flips
  
+ The probability mass function of a binomial random variable with success probability $p$ and $n$ trials is given by

. . .

$$
f(x) = P(X = x) = \binom{n}{x} p^x (1-p)^{n-x}
$$

::: {.no-bullet}
where $\binom{n}{x}$ is the binomial coefficient, which represents the number of ways to choose $x$ successes from $n$ trials
:::

## binomial distribution

+ the cdf of a binomial random variable is given by

. . .

$$
F(x) = P(X ≤ x) = \sum_{k=0}^{x} \binom{n}{k} p^k (1-p)^{n-k}
$$

::: {.no-bullet}
for $x = 0, 1, ..., n$
:::

+ use `dbinom()` to compute the probability of getting exactly $2$ heads in $10$ flips of a fair coin

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the probability of getting 2 heads in 10 flips of a fair coin"

dbinom(2, size = 10, prob = 0.5) # probability of getting 2 heads in 10 flips
```


+ the cdf of a binomial random variable can be computed using the `pbinom()` function

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the probability of getting at most 2 heads in 10 flips"

pbinom(2, size = 10, prob = 0.5) # probability of getting at most 2 heads in 10 flips
```

  
::: notes

+ a binomial random variable is the sum of n independently and identically distributed (or i.i.d. in short) Bernoulli random variables 
+ Recall that a Bernoulli random variable equals either 1 or 0 with success probability p
  + Thus, X can take an integer value from 0 to n

:::


## Normal Distribution

+ A normal distribution is a continuous probability distribution that is `symmetric` about its mean, $\mu$, and characterized by its standard deviation, $\sigma$
  + The normal distribution is often referred to as the `Gaussian distribution`
 
+ The probability density function (pdf) of a normal distribution is given by

. . .

$$
f(x) = \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{(x - \mu)^2}{2\sigma^2}} \quad \textrm{for } -\infty < x < \infty
$$

+ The cumulative distribution function (cdf) of a normal distribution is given by

. . .

$$
F(x) = P(X ≤ x) = \int_{-\infty}^{x} f(t) dt \quad \textrm{for } -\infty < x < \infty
$$

+ $X \sim N(\mu, \sigma^2)$; when $\mu=0$ and $\sigma^2=1$, $X$ is said to follow **standard normal distribution**

::: notes

+ notice that the cdf of a normal distribution does not have a closed-form expression, but it can be computed using numerical methods

::: 

## Normal Distribution

+ The normal distribution $X \sim N(\mu, \sigma^2)$ has two important properties. Let $c$ be a constant
  1. let $Z=X+c$, then $Z \sim N(\mu+c, \sigma^2)$
  2. let $Z=cX$, then $Z \sim N(c\mu, c^2\sigma^2)$
  
+ `z-score` of a normal random variable $X$ is defined as

. . .

$$
\textrm{z-score} = \frac{X - \mu}{\sigma} \sim N(0,1)
$$

## Normal Distribution

+ note that when the data are distributed according to the normal distribution
  + about `two-thirds` are within $1$ standard deviation from the mean
  + approximately `95%` are within the $2$ standard deviations from the mean
  
. . .

$$
P(\mu - k\sigma \leq X \leq \mu + k\sigma) \approx \begin{cases}
0.68 & \textrm{if } k = 1 \\
0.95 & \textrm{if } k = 2 \\
0.997 & \textrm{if } k = 3
\end{cases}
$$

+ we can express the above probabilities in terms of `z-score`

. . .

$$
P(\mu - k\sigma \leq X \leq \mu + k\sigma) = P(-k \leq Z \leq k)=F(k) - F(-k)
$$

## Normal Distribution

+ use the `pnorm()` function in R to compute the cumulative distribution function of a normal distribution


```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the probability of being within 1 SD from $N(0, 1)$ using pnorm()"

pnorm(1) - pnorm(-1) # probability of being within 1 standard deviation from the mean

```

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the probability of being within 2 SDs from $N(0, 1)$ using pnorm()"


pnorm(2) - pnorm(-2) # probability of being within 2 standard deviations from the mean

```

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the probability of being within 3 SDs from $N(0, 1)$ using pnorm()"


pnorm(3) - pnorm(-3) # probability of being within 3 standard deviations from the mean

```

+ we can directly specify mean and standard deviation 

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the probability of being within 1 SD from $N(5, 2^2)$ using pnorm()"
#| 
mu <- 5
sigma <- 2

pnorm(mu + sigma, mean = mu, sd = sigma) -
  pnorm(mu - sigma, mean = mu, sd = sigma)

```

```{r}
#| echo: True
#| output-location: fragment
#| class: fragment
#| code-summary: "Compute the probability of being within 2 SDs from $N(5, 2^2)$ using pnorm()"

pnorm(mu + 2 * sigma, mean = mu, sd = sigma) -
  pnorm(mu - 2 * sigma, mean = mu, sd = sigma)

```

## Expectation

+ we used means and standard deviations in order to measure the `center` and `spread` of a distribution

+ Sample mean refers to the average of a variable in a particular data set

+ the expectation (or population mean) represents the mean value under a probability distribution

+ Sample mean fluctuates from one sample to another
  + but the expectation of a random variable is of a theoretical nature and is fixed given a probability model

:::: {.columns}

::: {.column width="50%"}

+ When $X$ is discrete

$$
E(X) = \sum_{x} x \cdot P(X = x)
$$

::: {.no-bullet}
where $P(X = x)$ is the probability mass function of the random variable $X$
:::

:::

::: {.column width="50%"}

+ When $X$ is continuous

$$
E(X) = \int_{-\infty}^{\infty} x \cdot f(x) dx
$$

::: {.no-bullet}
where $f(x)$ is the probability density function of the random variable $X$
:::

:::
::::


## Expectation - some examples

+ Consider a **Bernoulli** random variable $X$ with success probability $p$
  + The expectation of a Bernoulli random variable is given by
  
. . .

$$
E(X) = 0 \cdot (1-p) + 1 \cdot p = p
$$

+ Consider a **binomial** random variable $X$ with success probability $p$ and $n$ trials
  + The expectation of a binomial random variable is given by

. . .

$$
E(X) = \sum_{x=0}^{n} x \cdot P(X = x) = \sum_{x=0}^{n} x \cdot \binom{n}{x} p^x (1-p)^{n-x} = n \cdot p
$$


## Expectation - some examples

+ Consider a **uniform** random variable $X$ with the interval $[a, b]$
  + The expectation of a uniform random variable is given by

. . .

$$
E(X) = \int_{a}^{b} x \cdot f(x) dx = \int_{a}^{b} x \cdot \frac{1}{b-a} dx = \frac{a + b}{2}
$$


## Expectation - some properties

+ Let $X$ and $Y$ be random variables and $a$ and $b$ be arbitrary constants

+ The expectation is a linear operator that satisfies the following equality

. . .

$$
E(aX + bY) = aE(X) + bE(Y)
$$


+ Example - the expectation of a binomial random variable $X$ with success probability $p$ and $n$ trials 

. . .

$$
E(X) = E\left(\sum_{i=1}^{n} Y_i\right) = \sum_{i=1}^{n} E(Y_i) = nE(Y_1) = np
$$

::: {.no-bullet}
where $Y_i$ is a Bernoulli random variable with success probability $p$ and $E(Y_i) = p$
:::

<!--

+ A key result: if $X$ and $Y$ are independent random variables, then

. . .

$$
E(XY) = E(X)E(Y)
$$
-->

## Variance

+ The variance of a random variable $X$ is a measure of the spread of the distribution of $X$ around its expectation

+ The variance of a random variable $X$ is defined as

. . .

$$
\textrm{Var}(X) = E((X - E(X))^2) = E(X^2) - (E(X))^2
$$

+ the expectation of $X^2$ is called the *second moment* of the random variable $X$

+ the square root of the variance is called the **standard deviation** of the random variable $X$

+ example
  + the variance of a Bernoulli random variable $X$ with success probability $p$ is given by
  
. . .

$$
\textrm{Var}(X) = E(X^2) - (E(X))^2 = p(1-p) - p^2 = p(1-p)
$$

## Variance - properties

+ Let $X$ and $Y$ be random variables and $a$ and $b$ be arbitrary constants

+ The variance operator $V$ has the following properties
  + $V(a)=0$
  + $V(X+b) = V(X)$
  + $V(aX) = a^2V(X)$

+ if $X$ and $Y$ are independent, $V(aX + bY) = a^2V(X) + b^2V(Y)$

+ Example 1
  + the variance of a binomial random variable $X$ with success probability $p$ and $n$ trials is given by
  
. . .

$$
V(X) = V\left(\sum_{i=1}^{n} Y_i\right) = \sum_{i=1}^{n} V(Y_i) = nV(Y_1) = np(1-p)
$$

::: {.no-bullet-nested}
where $Y_i$ is a Bernoulli random variable with success probability $p$ and $V(Y_i) = p(1-p)$
:::

<!--

+ Example 2
  + the variance of the sum of two independent normal random variable $X \sim N(\mu_X, \sigma^2_X)$ and $Y \sim N(\mu_Y, \sigma^2_Y)$ can be computed as

. . .

$$
V(X + Y) = V(X) + V(Y) = \sigma^2_X + \sigma^2_Y
$$
-->

## Example - Predicting election outcomes with uncertainty

+ the prediction of election outcomes using pre-election polls
  + the prediction of election outcomes is often based on pre-election polls, which are surveys conducted to estimate the preferences of voters before an election
  
+ polling has sampling variability because we only interview a fraction of a large population

+ Let us focus our discussion on the `US 2008 election`
  + Suppose that the Election Day outcome represents the true proportion of `Obama` and `McCain` supporters in the population of voters within each state and we assume that the fraction of voters who support a third-party candidate is negligible
    + the two-party support rate for Obama $p_j$, and McCain $1 − p_j$, within each state $j$ 

+ The CSV data file **pres08.csv** contains the 2008 US presidential election result

. . .


```{r}

library(tidyverse)

pres08 <- read_csv(
  "https://raw.githubusercontent.com/kwan-MSDA/soci7016_2025/refs/heads/main/data/pres08.csv"
)


#glimpse(pres08)

pres08 <- pres08 %>%
  mutate(p = Obama / (Obama + McCain))

```

::: notes

+ Suppose that we conduct a preelection poll under exactly the same conditions multiple times
  + Each time, we obtain a representative sample of the target population and yet the sample consists of different voters
  + This means that the estimated support for a candidate will differ for each sample
  
:::

## Predicting election outcomes with uncertainty

### MC simulation

+ assume that for each hypothetical sampling, we will interview $1000$ voters who are randomly selected from the population

+ The binomial distribution with success probability $p$ and size $1000$ within each state will model Obama’s support estimate
  + Using Monte Carlo simulation, we estimate Obama’s support within each state, then allocate that state’s Electoral College votes to the winning candidate
  + repeat this procedure **many times** to describe the uncertainty in pre-election polling estimates that is due to sampling variability
  + use the `rbinom()` function in $R$ and the prob argument of this function can take a vector of success probabilities

```{r}
#| class: fragment
#| code-summary: "Simulate the election outcome using Monte Carlo simulation"

set.seed(123)

sim_election <- function(.id, df, n_draws = 1000) {
  ## for each state randomly sample
  mutate(df, draws = rbinom(n(), n_draws, p)) %>%
    filter(draws > (n_draws / 2)) %>%
    summarize(EV = sum(EV), .id = .id)
}
```

## Predicting election outcomes with uncertainty

### MC simulation

+ use this function to simulate the election $10000$ times
  + use the `map_df()` function from the `purrr` package to apply an action multiple times

+ plot the results as a histogram - a reference line for Obama’s actual number of electoral votes $(364)$

```{r}
#| class: fragment
#| code-summary: "Simulate the election outcome 10000 times"
#| output-location: column-fragment
#| out-height: 5in

library(purrr)

sims <- 10000

sim_results <- map_df(seq_len(sims), ~ sim_election(.x, pres08, n_draws = 1000))

stats <- sim_results %>%
  select(EV) %>%
  summarize_all(list(mean = mean, median = median, var = var, sd = sd))

# make the stats into the plot

ggplot(sim_results, aes(x = EV, y = ..density..)) +
  geom_histogram(bins = 12, fill = "lightblue", color = "black") +
  geom_vline(
    xintercept = 364,
    linetype = "dashed",
    linewidth = 0.5,
    color = "red"
  ) +
    # annotate the actual number of electoral votes
  annotate(
    "text",
    x = 364 + 5,
    y = 0.03,
    label = "Actual: 364",
    color = "red"
  ) +
    # add the mean and median lines
  geom_vline(
    xintercept = stats$mean,
    linetype = "dashed",
    linewidth = 0.5,
    color = "blue"
  ) +
    annotate(
    "text",
    x = stats$mean + 5,
    y = 0.01,
    label = paste("Mean:", round(stats$mean, 2)),
    color = "blue"
  ) +
  geom_vline(
    xintercept = stats$median,
    linetype = "dashed",
    linewidth = 0.5,
    color = "grey20"
  ) +
  annotate(
    "text",
    x = stats$median + 5,
    y = 0.02,
    label = paste("Median:", round(stats$median, 2)),
    color = "grey20"
  ) +

  labs(
    title = "Distribution of Electoral Votes for Obama",
    x = "Obama’s Electoral College votes",
    y = "Density"
  ) +
  theme_minimal()
```


::: notes

+ all prediction draws are above the winning threshold of 270 votes

+ While the highest density of the histogram roughly corresponds to the actual number of Electoral College votes Obama won, the distribution of predictions is skewed

:::

## Predicting election outcomes with uncertainty

+ We can also compute the expected value of Obama’s Electoral College votes using the probability model
  + The expected value is the average of all possible outcomes, weighted by their probabilities

+ Let $S_j$ represent the number of respondents (among a total of $1000$ respondents) for a pre-election poll who express support for Obama in state $j$

+ We use $v_j$ to denote the number of Electoral College votes for state $j$
  + the expected number of Obama’s Electoral College votes is
  
. . .

$$
E(\textrm{Obama's votes}) = \sum_{j=1}^{51} E(\textrm{Obama wins state j}) \cdot v_j = \sum_{j=1}^{51} P(S_j > 500) \cdot v_j
$$

::: {.no-bullet-nested}
where $P(S_j > 500)$ is the probability that Obama wins state $j$ in the pre-election poll
:::

## Predicting election outcomes with uncertainty

+ the variance of Obama’s Electoral College votes can be computed as

. . .

$$
\begin{align*}
V(\textrm{Obama's predicted votes}) &= \sum_{j=1}^{51} V(v_j \mathbb{I}\{S_j > 500\}) \\
& = \sum_{j=1}^{51} P(S_j > 500)(1 - P(S_j > 500)) \cdot v_j^2
\end{align*}
$$

::: {.no-bullet}
where $\mathbb{I}\{S_j > 500\}$ is an indicator function that equals $1$ if Obama wins state $j$ and $0$ otherwise

:::

```{r}
#| class: fragment
#| code-summary: "Compute the expected value and variance of Obama's Electoral College votes using pbinom()"
#| output-location: fragment

n_draws <- 1000

library(gt)
library(tidyverse)
library(tidyverse)

pres08 %>%
  mutate(
    pb = pbinom(n_draws / 2, size = n_draws, prob = p, lower.tail = FALSE)
  ) %>%
  summarize(
    mean = sum(pb * EV),
    median = sum(EV * (pb > 0.5)),
    var = sum(pb * (1 - pb) * EV^2),
    sd = sqrt(var)
  ) %>% 
    gt() %>% 
    tab_header(
      title = "Expected Value and Variance of Obama's Electoral College Votes"
    ) %>%
    tab_source_note(
      source_note = "Note: The expected value and variance are computed using the probability model."
    )

```


## References {visibility="uncounted"}

::: {#refs}
:::

    