---
title: "R - An Introduction"
subtitle: "MSDA - Bootcamp 2025 Summer"
author: "KT Wong"
email: "kwanto@hku.hk"
affiliation: "Faculty of Social Sciences, HKU"

date: last-modified
date-format: medium

abstract: |
    The materials in this topic are drawn from @QSS, @Wickham2023, @Wickham2019 and @Wickham2016 as well as other sources, including Princeton Sociology Methods Camp 2023. The materials are for educational purposes only.
 


bibliography: reference_bootcamp.bib
format:
  revealjs:
    theme: clean_42_rb_revised5.scss
    self-contained-math: true
    embed-resources: true
    show-notes: false
    multiplex: false
    slide-number: c/t
    preview-links: auto
    margin: 0.03
    height: 1080
    width: 1920
    max-scale: 2
    smaller: false
    scrollable: true
    incremental: true
    template-partials:
      - title-slide.html
    controls: true
    touch: false
    code-fold: true
    code-overflow: scroll
    code-line-numbers: true
      
execute:
  echo: false
  eval: false
  warning: false
  message: false
  error: false
  cache: false
  

jupyter: python3
engine: knitr

knitr:
  opts_chunk:
    echo: true
fig-cap-location: bottom
fig-width: 9
fig-height: 5
#fig-dpi: 300
lightbox: auto
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(tidyverse)

```

## Introduction

+   R is a programming language and software environment for statistical computing and graphics
    +   widely used in academia and industry
    +   for data analysis, statistical modeling, and visualization
    +   for machine learning, data mining, and big data analysis
    +   for reproducible research and scientific computing
    
    
+   R is an implementation of the S programming language
    + S was developed at Bell Laboratories by John Chambers and colleagues
    + S was designed for data analysis and graphics
    + S was the precursor to the commercial statistical software package S-PLUS

## Introduction

+   R was created by Ross Ihaka and Robert Gentleman at the University of Auckland, New Zealand
    + named partly after the first names of the first two R authors and partly as a play on the name of S
    + a GNU project,  a free software movement initiated by Richard Stallman in 1983
    + freely available under the GNU General Public License
    + a large number of packages available

## Introduction

+ [tidyverse](https://www.tidyverse.org/) is a collection of R packages designed for data science
    + tidyverse packages share an underlying design philosophy, grammar, and data structures
    + tidyverse packages are designed to work together

+  focus on the tidyverse
    + mainly because it is easier to understand
    + do most data manipulation we need in social science research with these tools
    
+   A good reference for tidyverse is the book by @Wickham2019
    + [R for Data Science](https://r4ds.had.co.nz/)

## Introduction

+   for quick reference, visit Posit Primers on [Data Science](https://rstudio.cloud/learn/primers)
      + R Basics
      + Transform Tables
      + Visualize Data
  
+  for more specific topics, visit [Data Science](https://rstudio.cloud/learn/intermediate)
      + Data Wrangling
      + Data Visualization
      + Modelling

+   we use cover some base R functions first, then move on to the tidyverse

## Basic Arithmetic

+ R can be regarded as a calculator
  + it recognizes all of the standard arithmetic operators

. . .

```{r}
#| echo: true

# Basic arithmetic operations in R

1+2 

6-7 

5/2 

2^3

2+5*2^3

3*(4+5) 

(3+4)*5

sqrt(4)

```


+ modulo operators (integer division & remainder)

. . .

```{r}
#| echo: true

# Modulo operators in R

5 %/% 2  # integer division
5 %% 2   # remainder

100 %/% 60 ## How many whole hours in 100 minutes?

100 %% 60 ## How many minutes are left over?


```

::: notes

+ In RStudio, you can start a new script using the dropdown menu File > New File > R Script

+ use your R script to save finalized commands
  + type in the console to practice and debug, then use the script to save what worked
  
+ R scripts allow you to save comments, or annotations, to your code
  +  anything following the # symbol in a script will not be treated like an R command—essentially, R will skip over it

:::


## Logic

+ R can also be used for logical operations
  + it recognizes the standard logical operators

. . .

```{r}
#| echo: true

1 > 2

1 > 2 & 1 > 0.5 # The "&" stands for "and"

1 > 2 | 1 > 0.5 # The "|" stands for "or" (not a pipe a la the shell)

isTRUE (1 < 2)

```

+ Here are the [detailed list](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Logic.html) of logical operators in R
  + `&` and `|` are used for element-wise logical operations
  + `&&` and `||` are used for short-circuit logical operations
    + e.g. `1 > 2 && 1 > 0.5` will return `FALSE` without evaluating the second condition
  + `!` is used for negation (i.e. "not")
  + `==`, `!=`, `<`, `<=`, `>`, and `>=` are used for comparison

+ Here is further information about logical vectors in R
  + [Logical vectors](https://stat.ethz.ch/R-manual/R-devel/library/base/html/logical.html) are vectors that contain only `TRUE`, `FALSE`, or `NA` values
  + Logical vectors are often used for subsetting data frames or matrices

## Logic

+ Order of precedence
  + logic statements follow a strict order of precedence
  + logical operators (`>`, `==`, etc) are evaluated before Boolean operators (`&` and `|`)
    + failure to recognise this can lead to unexpected behaviour

. . .

```{r}

1 > 0.5 & 2

```

. . .

```{r}
#| echo: true

1 > 0.5 & 1 > 2


```

+ Lesson: explicit about each component of your logic statement(s)

::: notes

+ 2 is converted into `as.logical(2)` 

+ `as.logical(2)` is `TRUE` because it is not zero

:::

+ Negation: `!`
  + this will come in very handy when we start filtering data objects
  
. . .

```{r}
#| echo: true

is.na(1:10)
!is.na(1:10)

Negate(is.na)(1:10)

```

## logical operators

+ Value matching: `%in%`
   + whether an object is contained within (i.e. matches one of) a list of items, use `%in%`
   
. . .

```{r}
#| echo: True

4 %in% 1:10

4 %in% 5:10

```


+ There's no equivalent "not in" command
  + but we can create one

. . .

```{r}

`%ni%` = Negate(`%in%`) ## The backticks (`) help to specify functions.

4 %ni% 5:10

```



+ Evaluation
  + use two equal signs for logical evaluation

. . .

```{r}
#| echo: true

1 == 1 ## This does.

1 != 2 ## Note the single equal sign when combined with a negation

```

::: notes

+ Evaluation caveat: Floating-point numbers

+ What do you think will happen if we evaluate `0.1 + 0.2 == 0.3`?

+ 0.1 + 0.2 == 0.3

+ **Problem:** Computers represent numbers as binary (i.e. base 2) floating-points
  + More [here](https://floating-point-gui.de/basic/).
    + Fast and memory efficient, but can lead to unexpected behaviour. 

+ Similar to the way that standard decimal (i.e. base 10) representation can't precisely capture certain fractions
  + e.g. $\frac{1}{3} = 0.3333...$

+ **Solution:** Use `all.equal()` for evaluating floats (i.e  fractions)
  + all.equal(0.1 + 0.2, 0.3)

:::


## Assignment operators

+ Assignment: `<-` or `=`
  + use `<-` for assignment, but `=` is also acceptable
  + `<-` is the preferred operator in R, but `=` is more common in other programming languages

. . .

```{r}
#| echo: true

a <- 10 + 5
a
```

+ nevertheless, we can use `=` for assignment

. . .

```{r}
#| echo: true

b = 10 + 10 

b
```

::: notes

+ an arrow can point in the other direction too (i.e. `->`)

+ the following code chunk is equivalent to the previous one, although used much less frequently

```{r}
#| echo: true

10 + 5 -> a

```

+ Note that the assigned object *must* be on the left with "="

:::


## Help

+ Use `help` or `?` to get help on a function
  + e.g. `?mean` will give you the documentation for the `mean()` function
  
+ For more information on a (named) function or object in R, consult the "help" documentation. For example:

. . .

```{r}
#| echo: true

help(plot)

?plot 

```

+ Comments in R are demarcated by `#`. 
- Hit `Ctrl+Shift+c` in RStudio to (un)comment whole sections of highlighted code


+ Vignettes
  + for many packages, you can also try the `vignette()` function, which will provide an introduction to a package and it's purpose through a series of helpful examples
    + e.g. `vignette("ggplot2")` will give you a detailed introduction to the `ggplot2` package

::: notes

+ see the *Examples* section at the bottom of the help file?
  + You can run them with the `example()` function
  + Try it: `example(plot)`
  
:::

## Object-oriented programming (OOP) in R

+ R is an object-oriented programming language
  + everything in R is an object
  + objects can be manipulated and transformed using functions
  + functions are also objects in R


+ R has three main object classes
  + vectors
  + lists
  + data frames
  
+ R also has other object classes, such as matrices, arrays, and factors

+ Each object class has its own set of rules ("methods") for determining valid operations
  + e.g. you can perform many of the same operations on matrices and data frames
  + But there are some operations that only work on a matrix, and vice versa
  + At the same time, you can (usually) convert an object from one type to another
  
. . .

```{r d}
#| echo: true

## Create a small data frame called "d".

d = data.frame(x = 1:2, y = 3:4) 

d

```

. . .

```{r m}
#| echo: true

## Convert it to (i.e. create) a matrix call "m".

m = as.matrix(d)

m

```

::: notes

+ R can store information as an object with a name of our choice
  +  Once we have created an object, we just refer to it by name
  + we are using objects as “shortcuts” to some piece of information or data

+ there are many different *types* (or *classes*) of objects

+ arithmetic operations like addition and subtraction cannot be used for character strings

+ Most obviously, there are actually _multiple_ OOP frameworks in R
  + **S3**, **S4**, **R6**...
  + Hadley Wickham's "Advanced R" provides a [very thorough overview](https://adv-r.hadley.nz/oo.html) of the main ones

+ But for our purposes, it is much more helpful to think about 
  (a) the shared characteristics of these different systems and 
  (b) the broad implications of OOP in R.
    +  What we lose in detail, we hopefully gain in perspective
    +  But do read Hadley's book if you get the chance
      + It's incredibly helpful (as are all his books)
  
:::

## More on matrices

[*`Matrices in R`*]{.big}

+   A matrix is a two-dimensional array
    +   all elements must be of the same type
    +   can be created using the matrix() function
    +   can be created from a vector using the dim() function

. . .

```{r create_matrix}
#| eval: false

matrix1<- matrix(1:15, nrow=3, ncol=5)

matrix1

```


## More on matrices

[*`Matrices in R - some basic operations`*]{.big}

. . .

```{r create_matrix2}
#| eval: false

dim(matrix1)

```

. . .

```{r create_matrix1}
#| eval: false

colnames(matrix1)<- c("A", "B", "C", "D", "E")
rownames(matrix1)<- c("X", "Y", "Z")

matrix1

```

. . .

```{r create_matrix3}
#| eval: false

matrix1[2,3]

```

. . .

```{r create_matrix4}
#| eval: false

matrix1[2,]


```

## More on matrices

[*`Matrices in R - some basic operations`*]{.big}

. . .

```{r create_matrix5}
#| echo: true
#| eval: false

A<- matrix(1:6, nrow=2, ncol=3)
B<- matrix(7:12, nrow=3, ncol=2)


print(A %*% B)

print(B %*% A)

```

. . .

```{r create_matrix6}
#| echo: true
#| eval: false

print(t(A))

```

## More on matrices

[*`Matrices in R - some basic operations`*]{.big}

. . . 

```{r create_matrix7}
#| eval: false

C<- matrix(c(2,5,3,1,3,6,2,9,5), nrow=3, ncol=3)
print(solve(C))

```

. . .

```{r create_matrix8}
#| eval: false

print(det(C))

```
. . .

```{r create_matrix9}
#| eval: false

print(eigen(C))

```



## Object class, type, and structure

+ Use the `class`, `typeof`, and `str` commands if you want understand more about a particular object

. . .

```{r}
#| echo: True

d = data.frame(x = 1:2, y = 3:4) 

class(d) ## Evaluate its class

typeof(d) ## Evaluate its type

str(d) ## Show its structure

```

+ we can inspect/print an object directly in the console
  + e.g. Type `d`


+ The `View()` function is also very helpful
  + this is the same as clicking on the object in the RStudio *Environment* pane
  + e.g. `View(d)`
  
::: notes

+ PS — Confused by the fact that `typeof(d)` returns "list"? See [here](https://stackoverflow.com/questions/45396538/typeofdata-frame-shows-list-in-r).

:::

## changing data types

+ we might change the data type of a variable for some data analysis
    +   as.character()
    +   as.numeric()
    +   as.factor()
    +   as.integer()
    +   as.logical()

```{r}
#| class: fragment

agevec<- c(24, 22, 25, 28, 23, 20)

class(agevec)

```

```{r}
#| class: fragment

gendervec <- c("male", "female", "other", "female", "female", "male")

gendervec

class(gendervec)

```

```{r}
#| class: fragment

a<- c(28, "28", TRUE)

class(a)

```

```{r}
#| class: fragment

b<- c(1,2,3, TRUE, FALSE)

class(b)

```
## changing data types

+ Vectors can have a factor type
    + looks like a character vector
    + but is actually a number under the hood (“labelled data”)

```{r}
#| class: fragment

genderfactorvec<- factor(gendervec,
                         levels=c("male", "female", "other"))

genderfactorvec

class(genderfactorvec)

```


```{r}
#| class: fragment

as.numeric(genderfactorvec)

```


## Global environment

+ The global environment is where all objects are stored
  + you can see the global environment in the RStudio *Environment* pane
  + you can also see the global environment by typing `ls()` in the console
  
+ Working with multiple objects
  + you can create multiple objects in the global environment
  + you can manipulate and transform these objects using functions
  + you can also use the `rm()` function to remove objects from the global environment
  
+ R's ability to keep multiple objects in memory at the same time is a huge plus when it comes to effective data work 
  +  e.g. We can copy an existing data frame, or create new one entirely from scratch - either will exist happily with our existing objects in the global environment
  
. . .

```{r}
#| echo: true

d2 = data.frame(x = rnorm(10), y = runif(10))

```

## Everything has a name

+ Every object in R has a name
  + you can use the name to refer to the object in the global environment
  + object names are case sensitive
  
+ The name of our object must follow certain restrictions
  + it cannot begin with a number
  + it should not contain spaces
  + special characters such as % and $ should not be used

+ Reserved words
  + we can assign objects to different names
  + However, there are a number of special words that are "reserved" in R
    + These are are fundamental commands, operators and relations in base R that you cannot (re)assign, even if you wanted to
    + See [here](http://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html) for a full list

. . .

```{r}
#| echo: false
#| eval: false
#| include: false


if 
else 
while 
function 
for
TRUE 
FALSE 
NULL 
Inf 
NaN 
NA 

```

::: notes

+ you can also use the name to refer to the object in a function

+ function itself is an object
  + e.g. `mean` is a function that calculates the mean of a vector
  + we can assign it to a new name, e.g. `average = mean`
  + but we cannot assign it to a reserved word, e.g. `if = mean`
  + this will lead to an error

:::


## Everything has a name

+ Semi-reserved words
  + there is a class of words and strings that are called "semi-reserved"

. . .

```{r}
#| echo: true
my_vector = c(1, 2, 5)
my_vector

```

+ what happens if you type the following?

. . .

```{r}
#| echo: true

c = 4
c(1, 2 ,5)

```

+ R is "smart" enough to distinguish between the variable `c = 4` that we created and the built-in function `c()` that calls for concatenation

+ However, R won't always be able to distinguish between conflicting definitions

. . .

```{r}
#| echo: true

pi
pi = 2
pi

```

+ **Bottom line:** Don't use (semi-)reserved characters!

::: notes

+ these are named functions or constants (e.g. `pi`) that you can re-assign if you really wanted to... but already come with important meanings from base R
+ The `c()` function is a good example of a semi-reserved character
+ Arguably the most important semi-reserved character is `c()`, which we use for concatenation
    + i.e. creating vectors and binding different objects together

:::


## Namespace conflicts

+ R has a large number of packages, each with its own set of functions
  + sometimes, different packages have functions with the same name
  + this can lead to "namespace conflicts" where R doesn't know which function to use
  
+ e.g. Look what happens we load the `dplyr` package

. . .

```{r}
#| echo: true

library(dplyr)

```


+ The messages that you see about some object being *masked from 'package:X'* are warning you about a namespace conflict
  + e.g. Both `dplyr` and the `stats` package have functions named "filter" and "lag"

+ Whenever a namespace conflict arises, the most recently loaded package will gain preference
  + so the `filter()` function now refers specifically to the `dplyr` variant
  + what if we want the `stats` variant? 

::: notes

+ The potential for namespace conflicts is a result of the OOP approach
  + Also reflects the fundamental open-source nature of R and the use of external packages
  + People are free to call their functions whatever they want, so some overlap is only to be expected

+ Temporarily use `stats::filter()`
  + Permanently assign `filter = stats::filter`

:::

## Namespace conflicts

Two common ways to resolve namespace conflicts in R:

1. Use `package::function()`
  + We can explicitly call a conflicted function from a particular package using the `package::function()` syntax

. . .

```{r}
#| echo: true

stats::filter(1:10, rep(1, 2))

```

. . .

```{r}
#| echo: true

dplyr::starwars ## Print the starwars data frame from the dplyr package

scales::comma(c(1000, 1000000)) ## Use the comma function, which comes from the scales package

```

  
2. Assign `function = package::function`
  + A more permanent solution is to assign a conflicted function name to a particular package
  + This will hold for the remainder of your current R session, or until you change it back
    
. . .

```{r, eval=F}
#| echo: true

filter = stats::filter ## Note the lack of parentheses
filter = dplyr::filter ## Change it back again

```

::: notes

+ + We can also use `::` for more than just conflicted cases
  + e.g. Being explicit about where a function (or dataset) comes from can help add clarity to our code
  + e.g. `dplyr::filter(iris, Species=="virginica")` is clearer than just `filter(iris, Species=="virginica")`

+ Another good rule of thumb is that you want to load your most important packages last
  + e.g. Load the tidyverse after you've already loaded any other packages

+ User-side namespace conflicts
  + A final thing to say about namespace conflicts is that they don't only arise from loading packages
  + They can arise when users create their own functions with a conflicting name
    + e.g. If I was naive enough to create a new function called `c()`
    
:::

## Indexing

+ Indexing is a way to access specific elements of an object
  + e.g. we can access specific elements of a vector, list, or data frame using indexing
  
+ There are two main ways to index objects in R
  + Option 1: `[]`
    + This is the most common way to index objects in R
    + It works for vectors, matrices, data frames, and lists
    + It can be used to access specific elements of an object, or to subset an object
  + Option 2: `$`
    + This is a special operator that is used to access specific elements of a list or data frame by name
    + It is often used to access columns of a data frame by name

::: notes

+ Indexing is done using square brackets `[]`

+ Indexing can be done using either numeric or logical vectors
  + Numeric indexing refers to the position of the element in the object
  + Logical indexing refers to a logical vector that indicates which elements to keep
  
+ Indexing can also be done using names
  + e.g. we can access specific elements of a data frame using the column names 

:::

## Indexing

+ use `[]` to index objects that we create in R

. . .

```{r}
#| echo: true

a = 1:10
a[4] ## Get the 4th element of object "a"
a[c(4, 6)] ## Get the 4th and 6th elements

```

+ It also works on larger arrays (vectors, matrices, data frames, and lists)

. . .

```{r}
#| echo: true

library(dplyr)
starwars[1, 1] ## Show the cell corresponding to the 1st row & 1st column of the data frame

```

+ What does `starwars[1:3, 1]` give you?

+ What does `starwars[-1, 1]` give you?

+ what if we want to get the 1st column of the data frame?

. . .

```{r}
#| echo: true
#| eval: false

starwars[, 1] ## Get the 1st column of the starwars data frame
```

+ we can use `$` to access the columns of a data frame by name

. . .

```{r}
#| echo: true
#| eval: false

starwars$name ## Get the "name" column of the starwars data frame

```



::: notes

+ What if you want to get the names of the columns in a data frame?
. . .

```{r}
#| echo: false
#| eval: false

names(starwars) ## Get the names of the columns in the starwars data frame

```


+ [1] Indexing in R begins at 1. Not 0 like some languages (e.g. Python and JavaScript)

:::


## Indexing

+ **lists** are a more complex type of array object in R
  + they can contain a random assortment of objects that don't share the same class
    + e.g. a list can contain a vector, a data frame, and a matrix
    + or even lists of lists
  + lists are very flexible and can be used to store any type of object

+ The relevance to indexing is that lists require two square brackets `[[]]` to index the parent list item and then the standard `[]` within that parent item
+ This is different from the single square brackets that we use to index vectors, matrices, and data frames
  + The double square brackets are used to extract the object itself, rather than a subset of the list

## Indexing

+ Let's look at an example of a list to see how it works in practice
  + We'll use the `my_list` object, which contains three objects: a character string, a numeric vector, and a data frame

+ As before, we have two options for indexing lists

+ Option 1: `[[ ]]`
  + use double square brackets to index a list
  + this will return the object itself, rather than a subset of the list
  + e.g. `my_list[[1]]` will return the first object in the list
  + we can also use `my_list[[2]][3]` to return the third element of the second object 

. . .

```{r my_list}
#| echo: true

my_list = list(a = "hello", b = c(1,2,3), c = data.frame(x = 1:5, y = 6:10))

my_list[[1]] 

my_list[["a"]]

my_list[[2]][3]

```

## Indexing

+ Option 2: `$` 
  + lists provide a nice segue to our other indexing operator: `$`
  + Let's continue with the `my_list` example

. . .

```{r}
#| echo: true

my_list

my_list$a

my_list$c$x
```

+ The `$` form of indexing also works (and in the manner that you probably expect) for other object types in R
  + In some cases, you can also combine the two index options

. . .

```{r}
#| echo: true

my_list$b[3]

starwars$name[1]

```


::: notes

+ note some key differences between the output from this example and that of our previous `starwars[1, 1]` example
  + What are they?
  + Hint: Apart from the visual cues, try wrapping each command in `str()`

+ The last thing that I want to say about `$` is that it provides another way to avoid the "object not found" problem that we ran into with our earlier regression example
  + e.g. `lm(y ~ x)` will not work, but `lm(d$y ~ d$x)` will work

. . .

```{r, error=T}
#| echo: true

d = data.frame(x = 1:10, y = 11:20)

lm(y ~ x) 

lm(d$y ~ d$x) 

```

:::

## cleaning up

+ Use `rm()` to remove objects from the global environment
  + e.g. `rm(a)` will remove the object `a` from the global environment

. . .

```{r}
a = "hello"
b = "world"
rm(a, b)
```

+ use `rm(list = ls())` to remove all objects in your working environment (except packages)
  + Better just to start a new R session
  
::: notes

+ but this is [frowned upon](https://www.tidyverse.org/articles/2017/12/workflow-vs-script/)

:::


+ Detaching packages is more complicated
  + there could be many cross-dependencies (i.e. one package depends on, and might even automatically load, another package)
  + e.g. `detach(package:dplyr)`

+ still, better just to restart your R session
  

## Removing plots

+ Use `dev.off()` to remove the current plot
  + e.g. `dev.off()` will remove the current plot from the plotting window
  
+ Use `graphics.off()` to remove all plots

. . .

```{r, eval=F}
#| echo: true

plot(1:10)
dev.off()
```

+ RStudio has convenient buttons for clearing your workspace environment and removing (individual) plots

## Data files: Loading and saving

+ R can read and write data files in various formats
  + e.g. CSV, Excel, SPSS, Stata, etc.
  + we will focus on CSV files for now

+ working directory
  + R will look for files in the current working directory
  + you can check the current working directory using `getwd()`
  + you can change the working directory using `setwd()`
  + e.g. `setwd("path/to/your/directory")`

## Data files: Loading and saving

+ go to [here](https://raw.githubusercontent.com/kwan-MSDA/MSDA-Bootcamp/main/dataset/UNpop.csv){download="UNpop.csv"} to download a sample dataset

+ use `read.csv()` to read a CSV file into R
  + e.g. `unpop <- read.csv("UNpop.csv")`
  + this will read the CSV file and create a data frame called `unpop`

. . .

```{r}
#| echo: true

unpop <- read.csv("Downloads/UNpop.csv")

```

. . .

```{r}
#| echo: true

class(unpop)

```


+ RData can be used to save and load R objects
  + e.g. `save(unpop, file = "unpop.RData")` will save the `unpop` data frame as an RData file
  + you can load the RData file using `load("unpop.RData")`

+ RData files are useful for saving and loading multiple objects at once

. . .

```{r}
#| echo: true

save(unpop, file = "unpop.RData")

```


::: notes

+ a data frame is a special type of list where each element is a vector of the same length
  + data frames are used to store tabular data, where each column represents a variable and each row represents an observation

+ R will save the entire workspace as .RData in the working directory without an explicit file name and automatically load it next time we launch  
    + This is not recommended practice because the .RData file is invisible to users of many operating systems and R will not tell us what objects are loaded unless we explicitly issue the ls() function

:::


## Data files: Loading and saving

+ useful functions for exploring data frames
  + `names()` to view the names of the columns in a data frame
  + `nrow()` to view the number of rows in a data frame
  + `ncol()` to view the number of columns in a data frame
  + `dim()` to view the dimensions of a data frame (number of rows and columns)
  + `str()` to view the structure of a data frame
  + `head()` to view the first few rows of a data frame
  + `tail()` to view the last few rows of a data frame
  

+ use `write.csv()` to write a data frame to a CSV file
  + e.g. `write.csv(unpop, "UNpop.csv")`
  + this will write the `unpop` data frame to a CSV file called `UNpop.csv`

## Data files: Loading and saving

+ load data in other formats

+ R can read and write data files in various formats, including SPSS, Stata, and Excel

+ `foreign` package can read and write data files in SPSS, Stata, and other formats
  + it is part of the base R installation and is widely used for reading and writing data files
+ `haven` package is designed to be fast and efficient
  + it can read and write data files in SPSS, Stata, and SAS formats
  + it is part of the tidyverse and is designed to be fast and efficient
+ `readr` package can read and write data files in CSV, TSV, and other text formats
  + it is part of the tidyverse and is designed to be fast and efficient
+ `readxl` package can read and write Excel files
+ `writexl` package can write data frames to Excel files

. . .

```{r}
#| echo: true

#install.packages("haven")

library("haven")

```

. . .

```{r}
#| echo: true

# Read Stata file
url_stata<-"https://github.com/kwan-MSDA/MSDA-Bootcamp/blob/main/dataset/UNpop.dta?raw=true"
unpop_stata<- read_dta(url_stata)

```


## References {visibility="uncounted"}

::: {#refs}
:::

